---
title: "Nuclei to transcript niche hex bins"
output: html_document
params:
  sample:
    label: "Sample Name"
    input: text
    value: A-05_post-treatment
---

The following code (parameterised report) relates to the processing and mapping of Xenium spatial transcriptomics nuclei to transcript niche hexbins (from GraphSAGE output).

Processing Xenium spatial transcriptomics sample:
`r params$sample`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(SingleCellExperiment)
library(scran)
library(ggplot2)
library(hexbin)
library(pheatmap)
library(RColorBrewer)
library(data.table)
library(readr)

transcript_niche_color_list <- c(`0` = "#9FCF87",
                                 `1` = "#F0A0FF", 
                                 `2` = "#9068AC",
                                 `3` = "#F77800",
                                 `4` = "#FFCC99",
                                 `5` = "#2BCE48",
                                 `6` = "#C39C94",
                                 `7` = "#005C31")
```

# Load data

```{r}
spe <- readRDS(paste0("/data/process/spe_", params$sample, "_processed.rds"))
cells <- fread(paste0("/data_processed/ST/", params$sample, "/cells.csv.gz"))
cells <- cells[cells$cell_id %in% colnames(spe), ]

bins <- read.csv(paste0("/output/graphsage/", params$sample, "_hexbin_info.csv"))
bins <- bins[bins$niche != "white", ]

summary(cells)
summary(bins)
```

# Avg transcripts per bin

```{r}
mean(bins$count, na.rm = TRUE)
```

# Niche percentages

```{r}
niche_percentage <- bins %>%
  group_by(niche) %>%
  summarise(count = n()) %>% 
  ungroup() %>%
  mutate(total_bins = sum(count),  
         percentage = (count / total_bins) * 100) 
niche_percentage 
```

# Assign cells to transcript-based niches

Assign based on euclidean distance of each nuclei centroid to the hex bin centroids (based on min distance)

```{r}
# uncomment for large samples (run functions in chunks)
#cells_to_bins <- function(cells, bins, chunk_size = 1000) {
#    if(nrow(cells) == 0 || nrow(bins) == 0) {
#        stop("Error: 'cells' or 'bins' data frame is empty.")
#    }
#
#    closest_bins_results <- data.frame(cell_index = integer(), closest_bin = integer())
#
#    for (start_idx in seq(1, nrow(cells), by = chunk_size)) {
#        end_idx <- min(start_idx + chunk_size - 1, nrow(cells))
#        cells_chunk <- cells[start_idx:end_idx, ]
#
#        distances <- expand.grid(cell_index = 1:nrow(cells_chunk), bin_index = 1:nrow(bins)) %>%
#            mutate(
#                cell_x = cells_chunk$x_centroid[cell_index],
#                cell_y = cells_chunk$y_centroid[cell_index],
#                bin_x = bins$x[bin_index],
#                bin_y = bins$y[bin_index],
#                distance = sqrt((cell_x - bin_x)^2 + (cell_y - bin_y)^2)
#            )
#
#        closest_bins_chunk <- distances %>%
#            group_by(cell_index) %>%
#            slice_min(order_by = distance, n = 1) %>%
#            ungroup() %>%
#            select(cell_index, closest_bin = bin_index)
#
#        closest_bins_chunk$cell_index <- closest_bins_chunk$cell_index + start_idx - 1
#        closest_bins_results <- rbind(closest_bins_results, closest_bins_chunk)
#    }
#
#    # Ensure closest_bins_results has an entry for every cell
#    if (nrow(closest_bins_results) < nrow(cells)) {
#        stop("Not all cells have a closest bin computed. Check your distance calculations and chunk logic.")
#    }
#
#    # Join and mutate to create closest_bin_id
#    cells <- cells %>%
#        mutate(cell_index = row_number()) %>%
#        left_join(closest_bins_results, by = "cell_index") %>%
#        mutate(closest_bin_id = bins$bin[closest_bin])
#
#    # Check for any missing values in the final column
#    if (anyNA(cells$closest_bin_id)) {
#        stop("Missing 'closest_bin_id' values after join. Check bin indices and bins data.")
#    }
#
#    select(cells, -closest_bin, -cell_index)
#}

#cells <- cells_to_bins(cells, bins, chunk_size = 1000)


cells_to_bins <- function(cells, bins) {
  browser()
  if(nrow(cells) == 0 || nrow(bins) == 0) {
    stop("Error: 'cells' or 'bins' data frame is empty.")
  }
  distances <- expand.grid(cell_index = 1:nrow(cells), bin_index = 1:nrow(bins))

  distances <- distances %>%
    mutate(
      cell_x = cells$x_centroid[cell_index],
      cell_y = cells$y_centroid[cell_index],
      bin_x = bins$x[bin_index],
      bin_y = bins$y[bin_index]
    )
  
  # Euclidean distance
  distances <- distances %>%
    mutate(distance = sqrt((cell_x - bin_x)^2 + (cell_y - bin_y)^2))
  
  # determine closest bins by finding the min distance for each cell
  closest_bins <- distances %>%
    group_by(cell_index) %>%
    slice_min(order_by = distance, n = 1) %>%
    ungroup() %>%
    select(cell_index, closest_bin = bin_index)
  
  cells <- cells %>%
    mutate(cell_index = row_number()) %>%
    left_join(closest_bins, by = "cell_index") %>%
    mutate(closest_bin_id = bins$bin[closest_bin]) %>%
    select(-closest_bin, -cell_index) 
  
  return(cells)
  
}

cells <- cells_to_bins(cells, bins)
write_csv(cells, file = paste0("/output/graphsage/niche_8_binwidth70/", params$sample, "_cells2niches.csv"))

gc()
```

# Prepare dataframe

```{r}
df_spe <- as.data.frame(colData(spe))
df_spe$annotation <- factor(df_spe$annotation)
df_spe$x <- cells$x_centroid[match(rownames(df_spe), cells$cell_id)]
df_spe$y <- cells$y_centroid[match(rownames(df_spe), cells$cell_id)]
df_spe$bin <- cells$closest_bin_id[match(rownames(df_spe), cells$cell_id)]
df_spe$niche <- bins$niche[match(df_spe$bin, bins$bin)]

annotation_proportions <- df_spe %>%
  group_by(annotation) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(total_count = sum(count, na.rm = TRUE), 
         proportion = (count / total_count) * 100)
annotation_proportions
```

# Avg cells per bin

```{r}
df_spe %>%
  group_by(bin) %>%
  summarise(cell_count = n()) %>%  
  summarise(average_cells = mean(cell_count)) 
```

# Neighbourhood enrichment

```{r}
res <- table(df_spe$niche, df_spe$annotation)
res_enrich <- res/chisq.test(res)$expected

annotations <- data.frame(Niche = as.factor(rownames(res_enrich)))
rownames(annotations) <- rownames(res_enrich)

plot <- pheatmap(t(res_enrich), color = c(rev(colorRampPalette(brewer.pal(9, "Blues"))(10)), 
                               colorRampPalette(brewer.pal(9, "YlOrRd"))(9)), 
           annotation_col = annotations,
         annotation_colors = list(Niche = transcript_niche_color_list),
         breaks=c(seq(0, 1, 0.1), seq(2, 5, 1), seq(6, 10, 1)))

plot

ggsave(paste0("/output/graphsage/", params$sample, "_heatmap.png"), plot = plot)
```

