---
title: "Differential Methylation Analysis of Methylation (EPIC v2) Array Data"
output: html_document
---

The following code relates to the analysis of all participant processed methylation array data 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(minfi)
library(parallel)
library(ComplexHeatmap) 
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(ggsci)
library(DMRcate)
library(methylKit)
library(missMethyl)
library(gridExtra)
library(clusterProfiler)
library(org.Hs.eg.db)
library(stringr)
library(RColorBrewer)
library(limma)
library(reshape2)
library(readxl)
```

## Load processed data

```{r}
mSetSqFlt <- readRDS(file="data/process/filtered_data.rds")
targets <- readRDS(file = "data/process/targets.rds")

# EPIC annotation 
epic <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
```

## M & Beta Values

```{r}
mVals <- getM(mSetSqFlt)
head(mVals)

bVals <- getBeta(mSetSqFlt)
head(bVals)

pal <- Polychrome::palette36.colors(20)
densityPlot(bVals, sampGroups=targets$Sample, main="Beta values", 
            legend=FALSE, xlab="Beta values", pal = pal)
legend("top", legend = levels(factor(targets$Sample)), 
       text.col=pal, cex = 0.5)
densityPlot(bVals, sampGroups=targets$Sample_Type, main="Beta values", 
            legend=FALSE, xlab="Beta values", pal = c("#0C7BDC", "#E66100"))
legend("top", legend = levels(factor(targets$Sample_Type)), 
       text.col=c("#0C7BDC", "#E66100"), cex = 0.8)
densityPlot(bVals, sampGroups=targets$Patient, main="Beta values", 
            legend=FALSE, xlab="Beta values", pal = pal)
legend("top", legend = levels(factor(targets$Patient)), 
       text.col=pal, cex = 0.8)
densityPlot(mVals, sampGroups=targets$Sample, main="M-values", 
            legend=FALSE, xlab="M values", pal = pal)
legend("topleft", legend = levels(factor(targets$Sample)), 
       text.col=pal, cex = 0.5)
```

one-sided Kolmogorov-Smirnov Test (Post-treatment vs Pre-treatment):

* Treated (Post-treatment) cells (First Group): testing if the methylation levels are stochastically lower in the IDHi treated cells.
* Untreated (Pre-treatment) cells (Second Group): This group serves as the baseline or comparison group.

This test assesses whether the treated group's methylation levels are consistently lower than those of the untreated group - finding a significant global decrease of DNA methylation in treated cells.

```{r}
# per patient density plots
get_stars <- function(p) {
  if (p < 0.001) {
    return("****")
  } else if (p < 0.01) {
    return("***")
  } else if (p < 0.05) {
    return("**")
  } else if (p < 0.1) {
    return("*")
  } else {
    return("ns")
  }
}

sig_df <- list()
for(i in unique(targets$Patient)){
    targets_subset <- targets[targets$Patient == i, ]
    targets_subset$Sample_Type <- as.factor(targets_subset$Sample_Type)
    
    bVals_subset <- bVals[, colnames(bVals) %in% targets_subset$Sample]
    
    density_plot <- densityPlot(bVals_subset, sampGroups=targets_subset$Sample_Type, main=paste("Beta values", i, sep = " "), 
            legend=FALSE, xlab="Beta values", pal = c("#0C7BDC", "#E66100"))
    legend("topright", legend = levels(targets_subset$Sample_Type), 
       text.col=c("#0C7BDC", "#E66100"), cex = 0.8)
    print(density_plot)
    
    pre-treat <- bVals_subset[, targets_subset$Sample_Type == levels(targets_subset$Sample_Type)[1], drop=FALSE]
    post-treat <- bVals_subset[, targets_subset$Sample_Type == levels(targets_subset$Sample_Type)[2], drop=FALSE]
    
        pre_vector <- as.numeric(pre-treat)
        post_vector <- as.numeric(post-treat)
        # one-sided Kolmogorov-Smirnov Test
        test_result <- ks.test(post_vector, pre_vector, alternative = "less")
        sig_df[[i]] <- test_result
        p <- test_result$p.value
        print(paste(i, "-", p))
        
        stars <- get_stars(test_result$p.value)
        p_value_text <- sprintf(stars, test_result$p.value)
        
        text(x = 0.5, y = max(density(pre_vector)$y, density(post_vector)$y), labels = p_value_text, cex = 1.2, col = "black", adj = c(0.5, 0.5))
}

for(i in unique(targets$Patient)){
    targets_subset <- targets[targets$Patient == i, ]
    targets_subset$Sample_Type <- as.factor(targets_subset$Sample_Type)
    
    bVals_subset <- bVals[, colnames(bVals) %in% targets_subset$Sample]
    
    svg(paste0("figures/beta_density_", i, ".svg"))
    density_plot <- densityPlot(bVals_subset, sampGroups=targets_subset$Sample_Type, main=paste("Beta values", i, sep = " "), 
            legend=FALSE, xlab="Beta values", pal = c("#0C7BDC", "#E66100"))
    legend("topright", legend = levels(targets_subset$Sample_Type), 
       text.col=c("#0C7BDC", "#E66100"), cex = 0.8)
    print(density_plot)
    dev.off()
}
```

## Probe-wise differential methylation (DMPs)

```{r}
sampleType <- factor(targets$Sample_Type)
individual <- factor(targets$Patient) 

design <- model.matrix(~0+sampleType+individual, data=targets)
colnames(design) <- c(levels(sampleType),levels(individual)[-1])

fit <- lmFit(mVals, design)
contMatrix <- makeContrasts(Pre-treatment-Post-treatment,
                           levels=design)
contMatrix

fit2 <- contrasts.fit(fit, contMatrix)
fit2 <- eBayes(fit2)

# DM CpGs at FDR < 0.05
summary(decideTests(fit2))
```
No significant up or downregulated differentially methylated CpGs (FDR < 0.05).  

```{r}
# results table of differentially methylated probes for Pre-treatment - Post-treatment contrast
epicSub <- epic[match(rownames(mVals),epic$Name),
                      c(1:4,12:19,22:ncol(epic))]
DMPs <- topTable(fit2, num=Inf, coef=1, genelist=epicSub)
head(DMPs)

# plot top 4 most significantly differentially methylated CpGs 
par(mfrow=c(2,2))
sapply(rownames(DMPs)[1:4], function(cpg){
  plotCpg(bVals, cpg=cpg, pheno=targets$Sample_Type, ylab = "Beta values")
})
```
