---
title: "Cell type annotation of Xenium spatial transcriptomics data"
output: html_document
---

The following code relates to the cell type annotation of Xenium spatial transcriptomics data.
Each sample was annotated separately using this same script (Sample A-05_pre-treatment used as an example here).

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(SeuratObject)
library(SingleR)
library(ggplot2)
library(SingleCellExperiment)
library(pals)
library(intrinsicDimension)
library(reticulate)
library(readxl)
library(viridis)
library(dplyr)
library(magrittr)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(bluster)
library(scater)
library(scuttle)
library(scran)
library(org.Hs.eg.db)
library(ggrepel)
library(stringi)

colours_anno <- c("#F28482", "#9D4EDD", "#6A994E", "#3A7CA5", "#EF233C",
               "#F9C8C7", "#CDDEB9", "#A7C957", "#63A2C8", "#000000", "#FFEA00", "#F4F3EE",
               "#FAA307", "#FFBA08", "#FFD60A", "#E85D04", "#F48C06", "#F2CC8F",
               "#656D4A", "#A4AC86", "#B6AD90", "#7F4F24", "#A68A64", "#D6D1C1",
               "#000000", "#ADB5BD",
               "#9D0208")

names(colours_anno) <- c("AC-like", "MES-like", "OPC-like", "NPC-like", "Progenitor",
                      "Astrocyte", "Oligodendrocyte", "OPC", "Neuron", "Vasculature", "Immune", "Unknown",
                      "T Cell", "NK Cell", "B Cell", "CD8 T Cell", "CD4 T Cell", "T Reg Cell", 
                      "Dendritic Cell", "Neutrophil", "Monocyte", "Microglia", "Macrophage", "Mast Cell",
                      "Endothelial", "Mural",
                      "Death")

## ruiz
colours_pop <- Polychrome::palette36.colors(20)
names(colours_pop) <- c("AC-like","Astrocyte","MES-like","Mural cell","Neuron","Oligodendrocyte","OPC","OPC-like","RG","Endothelial","NPC-like", "Mono", "TAM-BDM", "TAM-MG", "DC", "Mast", "CD4/CD8", "NK", "B Cell", "Plasma B")
## couturier
colours_popC <- Polychrome::palette36.colors(6)
names(colours_popC) <- c("Astro","Mesenchymal","Neuronal","Oligo","Progenitor","Unassigned")
```

# Load Data

```{r}
spe <- readRDS("/data/process/spe_A-05_pre-treatment_processed.rds")
```

# Clusters

```{r, echo=FALSE}
df <- reducedDim(spe, "UMAP")
df <- as.data.frame(df)
df$Sample <- spe$sample
df$Cluster <- spe$Cluster

colours_clusters <- Polychrome::palette36.colors(12)
names(colours_clusters) <- 1:12

ggplot(df[sample(1:nrow(df)),], aes(x=UMAP1, y=UMAP2, color=Cluster)) + geom_point(alpha=0.5) + 
  theme_classic() +
guides(colour = guide_legend(override.aes = list(alpha = 1, size=3))) + scale_color_manual(values=colours_clusters)
```

# Annotation

Automatic reference-based SingleR annotation

References used:

* Ruiz-Moreno et al. 2022
* Couturier et al. 2020
* Herring et al. 2022 (Lister)

Gene inputs used for annotation:

* Reference dataset HVGs

```{r}
# load reference datasets 
sc_gbm_big <- readRDS("/public_datasets/sc_gbm_big/sce_downsampled.rds")
couturier <- readRDS("/public_datasets/Deconvolution References/Couturier_dataset_tumor_lognormalised.rds")
reference <- readRDS("/public_datasets/lister/2020-12-18_whole-tissue_post-restaged-GABA-clustering.RDS")
reference$concat <- paste(reference$major_clust, reference$stage_ids, sep = "_")

# gene inputs
ReferenceHVGs <- function(x, symbol="feature_name"){
  rownames(x) <- rowData(x)[,symbol]
  dec <- modelGeneVar(x)
  Reference_HVGs <- getTopHVGs(dec, n=5000)
  return(Reference_HVGs)
}

reference.HVGs <- ReferenceHVGs(sc_gbm_big)
reference.HVGsC <- ReferenceHVGs(couturier, symbol = "Symbol")
reference.HVGsL <- ReferenceHVGs(reference, symbol = "index")

# function
annotate_clust <- function(sce, reference, genes, symbol="feature_name", annotation="annotation_level_3"){
  
  #Check for excessive amount of Ensembl gene names 
  is.ens <- grepl("^ENSG", genes, ignore.case = TRUE)
  num.true <- length(is.ens[is.ens == TRUE])
  if(num.true == length(genes)){
    print(warning("Ensembl genes detected in gene names"))
  }
  
  #Check Marker Genes are actually present in dataset of interest
  is.present <- genes %in% rownames(sce)
  num.false <- length(is.present[is.present == FALSE])
  if(num.false > 0){
    print(warning("Marker genes not present in dataset of interest. Possibly wrongly named"))
  }
  
  #Annotate clusters using reference dataset
  counts(reference) <- logcounts(reference)
  rownames(reference) <- rowData(reference)[,symbol]
  available <- intersect(rownames(reference), rownames(sce))
  pred_labels <- SingleR(test=sce[available,], ref=reference[available,], restrict = genes, 
                         labels=colData(reference)[,annotation], de.method="wilcox")
  return(pred_labels$pruned.labels)
}

# run
anno <- annotate_clust(sce = spe, reference = sc_gbm_big, genes = reference.HVGs)
annoC <- annotate_clust(sce = spe, reference = couturier, genes = reference.HVGsC, symbol = "Symbol", annotation = "cluster")
annoL <- annotate_clust(sce = spe, reference = reference, genes = reference.HVGsL, symbol = "index", annotation = "concat")

spe$anno_ruiz <- anno
spe$anno_cout <- annoC
spe$anno_lister <- annoL

gc()
```

```{r, echo=FALSE}
## ruiz
plotReducedDim(object = spe, dimred = "UMAP", colour_by = "anno_ruiz") + theme(aspect.ratio = 1)+ ggtitle("Ruiz-Moreno Annotation")+
  scale_color_manual(values=colours_pop)
## couturier
plotReducedDim(object = spe, dimred = "UMAP", colour_by = "anno_cout") + theme(aspect.ratio = 1)+ ggtitle("Couturier Annotation")+
  scale_color_manual(values=colours_popC)
```

```{r, echo=FALSE}
# Astro and Microglia and Oligo
spe_subset <- spe
astro_or_micro_cells <- grepl("^Astro", spe_subset$anno_lister) | grepl("^Micro", spe_subset$anno_lister) | grepl("^Oligo", spe_subset$anno_lister)
colData(spe_subset)$anno_lister[!astro_or_micro_cells] <- "Other_celltype"

plotReducedDim(object = spe_subset, dimred = "UMAP", colour_by = "anno_lister") + theme(aspect.ratio = 1)+ ggtitle("Lister Astro, Microglia & Oligo Cells")
```

## Inspect Marker Genes

```{r, include=FALSE}
# Inspect marker genes
markers <- function(x, marker){
  x$marker <- as.vector(logcounts(x[marker,]))
  plot <- plotReducedDim(object = x, dimred = "UMAP", colour_by = marker) + theme(aspect.ratio = 1)+ ggtitle(marker)+
  scale_color_viridis_c()
  return(plot)
}
```

### AC-like

```{r, echo=FALSE}
marker_genes <- c("APOE","AQP4", "MGAT4C")
lapply(marker_genes, function(marker) markers(spe, marker))
```

### MES-like

```{r, echo=FALSE}
marker_genes <- c("PDGFRA", "CHI3L1")
lapply(marker_genes, function(marker) markers(spe, marker))
```

### OPC-like

```{r, echo=FALSE}
marker_genes <- c("APOD", "OLIG1", "OLIG2")
lapply(marker_genes, function(marker) markers(spe, marker))
```

### Progenitor

```{r, echo=FALSE}
marker_genes <- c("TOP2A", "MKI67")
lapply(marker_genes, function(marker) markers(spe, marker))
```

### NPC-like

```{r, echo=FALSE}
marker_genes <- c("SOX11", "HES6", "STMN2")
lapply(marker_genes, function(marker) markers(spe, marker))
```

### Neuron

```{r}
marker_genes <- c("NPTXR", "STMN2", "CDH13", "C1QL3")
lapply(marker_genes, function(marker) markers(spe, marker))
```

### Vasculature

```{r, echo=FALSE}
marker_genes <- c("IGFBP7", "DCN")
lapply(marker_genes, function(marker) markers(spe, marker))
```

### Astrocyte

```{r, echo=FALSE}
marker_genes <- c("MGST1", "NTRK2")
lapply(marker_genes, function(marker) markers(spe, marker))
```

### Oligodendrocyte

```{r, echo=FALSE}
marker_genes <- c("ERMN", "MBP")
lapply(marker_genes, function(marker) markers(spe, marker))
```

### Immune Markers

```{r, echo=FALSE}
# Immune Markers

## T Cell
marker_genes <- c("CD4", "CD8A", "CCR7", "CD27", "PTPRC", "CD28", "KLRB1", "CD69", "FOXP3")
lapply(marker_genes, function(marker) markers(spe, marker))
## B-cells
marker_genes <- c("CD19", "CD79A", "BANK1")
lapply(marker_genes, function(marker) markers(spe, marker))
## Macrophage
marker_genes <- c("CD163", "CD68")
lapply(marker_genes, function(marker) markers(spe, marker))
## Neutrophil
marker_genes <- c("ITGAX", "SLC11A1", "CXCR1")
lapply(marker_genes, function(marker) markers(spe, marker))
## Monocyte
marker_genes <- c("VCAN", "FCER1G", "CD14")
lapply(marker_genes, function(marker) markers(spe, marker))
## Dendritic Cell
marker_genes <- c("FCER1A", "S100A4", "HLA-DQA1", "HLA-DQB1")
lapply(marker_genes, function(marker) markers(spe, marker))
## NK cells
marker_genes <- c("KLRC1", "GNLY") 
lapply(marker_genes, function(marker) markers(spe, marker))
## Microglia
marker_genes <- c("P2RY12", "CX3CR1")
lapply(marker_genes, function(marker) markers(spe, marker))
```

## Final Annotation

```{r, echo=FALSE}
# Final annotation
annotation_df <- read.csv(file = "/data/process/A-05_pre-treatment_annotations.csv")
spe$annotation <- annotation_df$annotation[match(spe$Cluster, annotation_df$cluster)]
spe$annotation <- factor(spe$annotation)
df$annotation <- spe$annotation

ggplot(df[sample(1:nrow(df)),], aes(x=UMAP1, y=UMAP2, color=annotation)) + geom_point(alpha=0.5) + 
  theme_classic() + labs(title = "A-05_pre-treatment")+
guides(colour = guide_legend(override.aes = list(alpha = 1, size=3))) + scale_color_manual(values=colours_anno) 

df %>%
  dplyr::count(annotation, Sample) %>%       
  dplyr::group_by(Sample) %>%
  dplyr::mutate(pct= prop.table(n) * 100) %>%
  ggplot(aes(y=Sample, x=pct, fill=annotation)) +
  geom_bar(stat="identity") +
  theme_bw()+ scale_fill_manual(values=colours_anno) 

#saveRDS(spe, file = "/data/process/spe_A-05_pre-treatment_processed.rds")
```

## Investigate/re-cluster NPC-like population

Are there normal neurons hiding here?

```{r}
spe_npc <- spe[, spe$annotation == "NPC-like"]

# re-cluster
set.seed(0010101010)
spe_npc <- runPCA(spe_npc, ncomponents=50,
    BSPARAM=BiocSingular::IrlbaParam(),  BPPARAM=MulticoreParam(10))

dim_est <- maxLikGlobalDimEst(reducedDim(spe_npc, "PCA"), k=20, unbiased=TRUE)
dim_est <- ceiling(dim_est$dim.est)

reducedDim(spe_npc, "PCA") <- reducedDim(spe_npc, "PCA")[,1:dim_est] 

set.seed(0010101010)
spe_npc <- runUMAP(spe_npc, dimred="PCA", name = 'UMAP',
   BPPARAM=MulticoreParam(10), spread = 1, min_dist = 0.1)

scater::plotReducedDim(spe_npc, dimred = "UMAP", color_by = "sample")

set.seed(0010101010)
clust.leiden <- clusterCells(spe_npc, use.dimred="PCA", 
    BLUSPARAM=SNNGraphParam(k=29, type="rank", cluster.fun="leiden"))
table(clust.leiden)
colData(spe_npc)$Cluster_npc <- clust.leiden

scater::plotReducedDim(spe_npc, dimred = "UMAP", color_by = "Cluster_npc")

# NPC-like
marker_genes <- c("SOX11", "HES6", "STMN2", "SOX4")
lapply(marker_genes, function(marker) markers(spe_npc, marker))

# Neuron
marker_genes <- c("NPTXR", "CDH13", "C1QL3", "EPHA4", "LMO4", "NRN1", "SPOCK3")
lapply(marker_genes, function(marker) markers(spe_npc, marker))

plotReducedDim(object = spe_npc, dimred = "UMAP", colour_by = "anno_cout") + theme(aspect.ratio = 1)+ ggtitle("Couturier Annotation")+
  scale_color_manual(values=colours_popC)
plotReducedDim(object = spe_npc, dimred = "UMAP", colour_by = "anno_ruiz") + theme(aspect.ratio = 1)+ ggtitle("Ruiz-Moreno Annotation")+
  scale_color_manual(values=colours_pop)

# Progenitor
marker_genes <- c("TOP2A", "MKI67")
lapply(marker_genes, function(marker) markers(spe_npc, marker))

annotation_df <- read.csv(file = "/data/process/A-05_pre-treatment_npc_anno.csv")
spe_npc$annotation <- annotation_df$annotation[match(spe_npc$Cluster_npc, annotation_df$cluster)]
spe_npc$annotation <- factor(spe_npc$annotation)
scater::plotReducedDim(spe_npc, dimred = "UMAP", color_by = "annotation")

# update annotations
levels(spe$annotation) <- union(levels(spe$annotation), unique(spe_npc$annotation))
colData(spe)$annotation[match(colnames(spe_npc), colnames(spe))] <- spe_npc$annotation
```

## Investigate/re-cluster AC-like population

Are there normal astrocytes hiding here?

```{r}
spe_ac <- spe[, spe$annotation == "AC-like"]

# re-cluster
set.seed(0010101010)
spe_ac <- runPCA(spe_ac, ncomponents=50,
    BSPARAM=BiocSingular::IrlbaParam(),  BPPARAM=MulticoreParam(10))

dim_est <- maxLikGlobalDimEst(reducedDim(spe_ac, "PCA"), k=20, unbiased=TRUE)
dim_est <- ceiling(dim_est$dim.est)

reducedDim(spe_ac, "PCA") <- reducedDim(spe_ac, "PCA")[,1:dim_est] 

set.seed(0010101010)
spe_ac <- runUMAP(spe_ac, dimred="PCA", name = 'UMAP',
   BPPARAM=MulticoreParam(10), spread = 1, min_dist = 0.1)

scater::plotReducedDim(spe_ac, dimred = "UMAP", color_by = "sample")

set.seed(0010101010)
clust.leiden <- clusterCells(spe_ac, use.dimred="PCA", 
    BLUSPARAM=SNNGraphParam(k=38, type="rank", cluster.fun="leiden"))
table(clust.leiden)
colData(spe_ac)$Cluster_ac <- clust.leiden

scater::plotReducedDim(spe_ac, dimred = "UMAP", color_by = "Cluster_ac")

# AC-like
marker_genes <- c("APOE","AQP4", "MGAT4C")
lapply(marker_genes, function(marker) markers(spe_ac, marker))

# Astrocyte
marker_genes <- c("MGST1", "NTRK2", "FGFR3", "GJA1", "TTYH1", "CST3")
lapply(marker_genes, function(marker) markers(spe_ac, marker))

plotReducedDim(object = spe_ac, dimred = "UMAP", colour_by = "anno_ruiz") + theme(aspect.ratio = 1)+ ggtitle("Ruiz-Moreno Annotation")+
  scale_color_manual(values=colours_pop)

# lister
colData(spe_ac)$anno_lister[!grepl("^Astro", spe_ac$anno_lister) ] <- "Other_celltype"
plotReducedDim(object = spe_ac, dimred = "UMAP", colour_by = "anno_lister") + theme(aspect.ratio = 1)

annotation_df <- read.csv(file = "/data/process/A-05_pre-treatment_ac_anno.csv")
spe_ac$annotation <- annotation_df$annotation[match(spe_ac$Cluster_ac, annotation_df$cluster)]
spe_ac$annotation <- factor(spe_ac$annotation)
scater::plotReducedDim(spe_ac, dimred = "UMAP", color_by = "annotation")

# update annotations
levels(spe$annotation) <- union(levels(spe$annotation), unique(spe_ac$annotation))
colData(spe)$annotation[match(colnames(spe_ac), colnames(spe))] <- spe_ac$annotation
```

## Investigate/re-cluster Immune population

Tease apart Immune sub-populations

```{r}
spe_im <- spe[, spe$annotation == "Immune"]

# re-cluster
set.seed(0010101010)
spe_im <- runPCA(spe_im, ncomponents=50,
    BSPARAM=BiocSingular::IrlbaParam(),  BPPARAM=MulticoreParam(10))

dim_est <- maxLikGlobalDimEst(reducedDim(spe_im, "PCA"), k=20, unbiased=TRUE)
dim_est <- ceiling(dim_est$dim.est)

reducedDim(spe_im, "PCA") <- reducedDim(spe_im, "PCA")[,1:dim_est] 

set.seed(0010101010)
spe_im <- runUMAP(spe_im, dimred="PCA", name = 'UMAP',
   BPPARAM=MulticoreParam(10), spread = 1, min_dist = 0.1)

scater::plotReducedDim(spe_im, dimred = "UMAP", color_by = "sample")

set.seed(0010101010)
clust.leiden <- clusterCells(spe_im, use.dimred="PCA", 
    BLUSPARAM=SNNGraphParam(k=21, type="rank", cluster.fun="leiden"))
table(clust.leiden)
colData(spe_im)$Cluster_im <- clust.leiden

scater::plotReducedDim(spe_im, dimred = "UMAP", color_by = "Cluster_im")

## T Cell
marker_genes <- c("CD4", "CD8A", "CCR7", "CD27", "PTPRC", "CD28", "KLRB1", "CD69", "FOXP3")
lapply(marker_genes, function(marker) markers(spe_im, marker))
## B-cells
marker_genes <- c("CD19", "CD79A", "BANK1")
lapply(marker_genes, function(marker) markers(spe_im, marker))
## Macrophage
marker_genes <- c("CD163", "CD68")
lapply(marker_genes, function(marker) markers(spe_im, marker))
## Neutrophil
marker_genes <- c("ITGAX", "SLC11A1", "CXCR1")
lapply(marker_genes, function(marker) markers(spe_im, marker))
## Monocyte
marker_genes <- c("VCAN", "FCER1G", "CD14")
lapply(marker_genes, function(marker) markers(spe_im, marker))
## Dendritic Cell
marker_genes <- c("FCER1A", "S100A4", "HLA-DQA1", "HLA-DQB1")
lapply(marker_genes, function(marker) markers(spe_im, marker))
## NK cells
marker_genes <- c("KLRC1", "GNLY") 
lapply(marker_genes, function(marker) markers(spe_im, marker))
## Microglia
marker_genes <- c("P2RY12", "CX3CR1")
lapply(marker_genes, function(marker) markers(spe_im, marker))

annotation_df <- read.csv(file = "/data/process/A-05_pre-treatment_immune_anno.csv")
spe_im$annotation <- annotation_df$annotation[match(spe_im$Cluster_im, annotation_df$cluster)]
spe_im$annotation <- factor(spe_im$annotation)
scater::plotReducedDim(spe_im, dimred = "UMAP", color_by = "annotation")

# update annotations
levels(spe$annotation) <- union(levels(spe$annotation), unique(spe_im$annotation))
colData(spe)$annotation[match(colnames(spe_im), colnames(spe))] <- spe_im$annotation
```

## Investigate/re-cluster Neuron population

Tease apart Neuron sub-populations

```{r}
spe_neuron <- spe[, spe$annotation == "Neuron"]

# re-cluster
set.seed(0010101010)
spe_neuron <- runPCA(spe_neuron, ncomponents=50,
    BSPARAM=BiocSingular::IrlbaParam(),  BPPARAM=MulticoreParam(10))

dim_est <- maxLikGlobalDimEst(reducedDim(spe_neuron, "PCA"), k=20, unbiased=TRUE)
dim_est <- ceiling(dim_est$dim.est)

reducedDim(spe_neuron, "PCA") <- reducedDim(spe_neuron, "PCA")[,1:dim_est] 

set.seed(0010101010)
spe_neuron <- runUMAP(spe_neuron, dimred="PCA", name = 'UMAP',
   BPPARAM=MulticoreParam(10), spread = 1, min_dist = 0.1)

scater::plotReducedDim(spe_neuron, dimred = "UMAP", color_by = "sample")

set.seed(0010101010)
clust.leiden <- clusterCells(spe_neuron, use.dimred="PCA", 
    BLUSPARAM=SNNGraphParam(k=21, type="rank", cluster.fun="leiden"))
table(clust.leiden)
colData(spe_neuron)$Cluster_n <- clust.leiden

scater::plotReducedDim(spe_neuron, dimred = "UMAP", color_by = "Cluster_n")

# Neuron
marker_genes <- c("NPTXR", "STMN2", "CDH13", "C1QL3", "EPHA4", "LMO4", "NRN1", "SPOCK3")
lapply(marker_genes, function(marker) markers(spe_neuron, marker))

genes <- read_xlsx("deconvolution_signatures/neuron_subtype_markers.xlsx")
genes <- genes[genes$gene %in% rownames(spe_neuron), ]

# Excitatory Neuron
marker_genes <- genes[genes$cell_type == "Excitatory Neuron", ]
marker_genes <- marker_genes$gene
lapply(marker_genes, function(marker) markers(spe_neuron, marker))

# Inhibitory Neuron
marker_genes <- genes[genes$cell_type == "Inhibitory Neuron", ]
marker_genes <- marker_genes$gene
lapply(marker_genes, function(marker) markers(spe_neuron, marker))

annotation_df <- read.csv(file = "/data/process/A-05_pre-treatment_neuron_anno.csv")
spe_neuron$annotation <- annotation_df$annotation[match(spe_neuron$Cluster_n, annotation_df$cluster)]
spe_neuron$annotation <- factor(spe_neuron$annotation)
scater::plotReducedDim(spe_neuron, dimred = "UMAP", color_by = "annotation")

# update annotations
levels(spe$annotation) <- union(levels(spe$annotation), unique(spe_neuron$annotation))
colData(spe)$annotation[match(colnames(spe_neuron), colnames(spe))] <- spe_neuron$annotation

scater::plotReducedDim(spe, dimred = "UMAP", color_by = "annotation")

## refine excitatory neurons
spe_neuron <- spe[, spe$annotation == "Excitatory Neuron"]

# re-cluster
set.seed(0010101010)
spe_neuron <- runPCA(spe_neuron, ncomponents=50,
    BSPARAM=BiocSingular::IrlbaParam(),  BPPARAM=MulticoreParam(10))

dim_est <- maxLikGlobalDimEst(reducedDim(spe_neuron, "PCA"), k=20, unbiased=TRUE)
dim_est <- ceiling(dim_est$dim.est)

reducedDim(spe_neuron, "PCA") <- reducedDim(spe_neuron, "PCA")[,1:dim_est] 

set.seed(0010101010)
spe_neuron <- runUMAP(spe_neuron, dimred="PCA", name = 'UMAP',
   BPPARAM=MulticoreParam(10), spread = 1, min_dist = 0.1)

scater::plotReducedDim(spe_neuron, dimred = "UMAP", color_by = "sample")

set.seed(0010101010)
clust.leiden <- clusterCells(spe_neuron, use.dimred="PCA", 
    BLUSPARAM=SNNGraphParam(k=15, type="rank", cluster.fun="leiden"))
table(clust.leiden)
colData(spe_neuron)$Cluster_n <- clust.leiden

scater::plotReducedDim(spe_neuron, dimred = "UMAP", color_by = "Cluster_n")

# Excitatory Neuron
marker_genes <- genes[genes$cell_type == "Excitatory Neuron", ]
marker_genes <- marker_genes$gene
lapply(marker_genes, function(marker) markers(spe_neuron, marker))

# Inhibitory Neuron
marker_genes <- genes[genes$cell_type == "Inhibitory Neuron", ]
marker_genes <- marker_genes$gene
lapply(marker_genes, function(marker) markers(spe_neuron, marker))

annotation_df <- read.csv(file = "/data/process/A-05_pre-treatment_excitatory_anno.csv")
spe_neuron$annotation <- annotation_df$annotation[match(spe_neuron$Cluster_n, annotation_df$cluster)]
spe_neuron$annotation <- factor(spe_neuron$annotation)
scater::plotReducedDim(spe_neuron, dimred = "UMAP", color_by = "annotation")

# update annotations
levels(spe$annotation) <- union(levels(spe$annotation), unique(spe_neuron$annotation))
colData(spe)$annotation[match(colnames(spe_neuron), colnames(spe))] <- spe_neuron$annotation

scater::plotReducedDim(spe, dimred = "UMAP", color_by = "annotation")
```

# Final cell type annotations

```{r}
df$annotation <- spe$annotation

ggplot(df[sample(1:nrow(df)),], aes(x=UMAP1, y=UMAP2, color=annotation)) + geom_point(alpha=0.5) + 
  theme_classic() + labs(title = "A-05_pre-treatment")+
guides(colour = guide_legend(override.aes = list(alpha = 1, size=3))) + scale_color_manual(values=colours_anno) 

saveRDS(spe, file = "/data/process/spe_A-05_pre-treatment_processed.rds")
```

