---
title: "Processing & quality control of Xenium spatial transcriptomics data"
output: html_document
params:
  sample: "sample name"
  neighbours:
    label: "Clustering Nearest Neighbours:"
    input: slider
    value: 40
    min: 30
    max: 80
---

The following code (parameterised report) relates to the processing and quality control of Xenium spatial transcriptomics data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(SeuratObject)
library(SingleR)
library(ggplot2)
library(SingleCellExperiment)
library(pals)
library(intrinsicDimension)
library(reticulate)
library(readxl)
library(viridis)
library(dplyr)
library(magrittr)
library(BiocParallel)
library(BiocNeighbors)
library(BiocSingular)
library(bluster)
library(scater)
library(scuttle)
library(scran)
library(org.Hs.eg.db)
library(ggrepel)
library(stringi)
```

Processing Xenium spatial transcriptomics sample - `r params$sample`

# Retain nuclei transcripts

```{r}
#load target sample
large_tran <- fread(paste0('/data_processed/ST/', params$sample, '/transcripts.csv.gz'), sep = ',')
smol_tran <- large_tran[large_tran$nucleus_distance == 0,]

#remove low quality transcripts (qv <20)
smol_tran <- smol_tran[as.numeric(smol_tran$qv)>=20,]

#turn into matrix
cell_list <- split(smol_tran$feature_name, smol_tran$cell_id)
cell_list <- lapply(lapply(cell_list,factor,levels = sort(unique(smol_tran$feature_name))),table)
cell_list <- lapply(cell_list, as.integer)
cell_matrix <- matrix(unlist(cell_list), ncol = length(cell_list), byrow = F)
colnames(cell_matrix) <- names(cell_list)
rownames(cell_matrix) <- sort(unique(smol_tran$feature_name))
sparse_mat <- as.sparse(cell_matrix)

#turn into SingleCellExperiment
spe <- SingleCellExperiment(assay = list('counts' = sparse_mat))

rm(large_tran)
rm(smol_tran)
rm(cell_matrix)
rm(cell_list)
gc()

saveRDS(spe, paste0('/data/tmp/spe_', params$sample, '.rds'))
```

# Load Xenium data

```{r}
spe <- readRDS(paste0("/data/tmp/spe_", params$sample, ".rds"))
spe$sample <- params$sample
```

# Filter & Normalise

```{r}
spe <- spe[,colSums(assay(spe, 'counts'))>20]
spe <- logNormCounts(spe, assay.type = 'counts')
```

# Dimension Reduction

```{r}
set.seed(0010101010)
spe <- runPCA(spe, ncomponents=50,
    BSPARAM=BiocSingular::IrlbaParam(),  BPPARAM=MulticoreParam(10))

dim_est <- maxLikGlobalDimEst(reducedDim(spe, "PCA"), k=20, unbiased=TRUE)
dim_est <- ceiling(dim_est$dim.est)
print(dim_est)

reducedDim(spe, "PCA") <- reducedDim(spe, "PCA")[,1:dim_est] 

set.seed(0010101010)
spe <- runUMAP(spe, dimred="PCA", name = 'UMAP',
   BPPARAM=MulticoreParam(10), spread = 1, min_dist = 0.1)

scater::plotReducedDim(spe, dimred = "UMAP", color_by = "sample")
```

## Clustering

* The number of nearest neighbours used was `r params$neighbours`

```{r}
set.seed(0010101010)
clust.leiden <- clusterCells(spe, use.dimred="PCA", 
    BLUSPARAM=SNNGraphParam(k=params$neighbours, type="rank", cluster.fun="leiden"))
table(clust.leiden)
colData(spe)$Cluster <- clust.leiden

scater::plotReducedDim(spe, dimred = "UMAP", color_by = "Cluster")
```

# Save output

```{r}
saveRDS(spe, file=paste0("/data/process/spe_", params$sample, "_processed.rds"))
```
