---
title: "Spitzer et al. 2024 comparison of snRNA-seq data"
output: html_document
---

The following code relates to the comparison of snRNA-seq data to data published in a similar study by Spitzer et al., 2024 (https://www.sciencedirect.com/science/article/pii/S153561082400093X?dgcid=author#sec4)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SummarizedExperiment)
library(HDF5Array)
library(SingleCellExperiment)
library(intrinsicDimension)
library(scater)
library(scuttle)
library(BiocParallel)
library(destiny)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(scran)
library(limma)
library(ggrepel)
library(tidyverse)
library(reshape2)
library(BiocParallel)
library(bluster)
library(fgsea)
library(pheatmap)
library(MASS)
library(lme4)
library(edgeR)
library(zoo)
library(Seurat)
library(GOSemSim)
library(RRHO2)

colours_comb <- c("#F28482", "#9D4EDD", "#6A994E", "#3A7CA5", "#EF233C",
               "#F9C8C7", "#CDDEB9", "#A7C957", "#63A2C8", "#000000", "#FFEA00", "#F4F3EE",
               "#FAA307", "#FFBA08", "#FFD60A", "#E85D04", "#F48C06", "#F2CC8F",
               "#656D4A", "#A4AC86", "#B6AD90", "#7F4F24", "#A68A64", "#D6D1C1",
               "#000000", "#ADB5BD",
               "#9D0208")
names(colours_comb) <- c("AC-like", "MES-like", "OPC-like", "NPC-like", "Progenitor",
                      "Astrocyte", "Oligodendrocyte", "OPC", "Neuron", "Vasculature", "Immune", "Unknown",
                      "T Cell", "NK Cell", "B Cell", "CD8 T Cell", "CD4 T Cell", "T Reg Cell", 
                      "Dendritic Cell", "Neutrophil", "Monocyte", "Microglia", "Macrophage", "Mast Cell",
                      "Endothelial", "Mural",
                      "Death")

colours_pop <- c("#F28482", "grey", "#000000", "#FFEA00", "#3A7CA5", "#6A994E", "#CDDEB9", "#FF474C", "#F4F3EE")
names(colours_pop) <- c("AC", "Doublet", "Endothel", "Macrophage", "NPC", "OC", "Oligodendrocyte", "Undifferentiated", "Unresolved")
```

# Annotate snRNA-seq data using Spitzer dataset as the reference

Automatic annotation run via SingleR

* Common HVGs between snRNA-seq data and Spitzer et al. reference (matched Oligodendroglioma dataset) used for SingleR scoring 

```{r}
sce_comb <- loadHDF5SummarizedExperiment("/data/process/", prefix="sce_integrated_")
# remove participant with low tumor purity
sce_comb <- sce_comb[, !sce_comb$Patient == "A-04"]
sce_comb$Patient <- factor(sce_comb$Patient)

# Spitzer SingleR results 
reference <- readRDS("/public_datasets/spitzer/GSE260928_SNRNA/matched_oligo_sce.rds")
# gene inputs
TopHVGs <- function(x){
  dec.parse <- modelGeneVar(x)
  TopHVGs <- getTopHVGs(dec.parse, n=5000)
  return(TopHVGs)
}
top.HVGs <- TopHVGs(sce)

ReferenceHVGs <- function(x, symbol="feature_name"){
  rownames(x) <- rowData(x)[,symbol]
  dec <- modelGeneVar(x)
  Reference_HVGs <- getTopHVGs(dec, n=5000)
  return(Reference_HVGs)
}
reference.HVGs <- ReferenceHVGs(reference, symbol = "Symbol")

CommonHVGs <- function(TopHVGs, ReferenceHVGs){
  CommonHVGs <- intersect(TopHVGs, ReferenceHVGs)
  return(CommonHVGs)
}
common.HVGs <- CommonHVGs(TopHVGs = top.HVGs, ReferenceHVGs = reference.HVGs)
length(common.HVGs)

# function
annotate_clust <- function(sce, reference, genes, symbol="feature_name", annotation="annotation_level_3"){
  
  #Check for excessive amount of Ensembl gene names 
  is.ens <- grepl("^ENSG", genes, ignore.case = TRUE)
  num.true <- length(is.ens[is.ens == TRUE])
  if(num.true == length(genes)){
    print(warning("Ensembl genes detected in gene names"))
  }
  
  #Check Marker Genes are actually present in dataset of interest
  is.present <- genes %in% rownames(sce)
  num.false <- length(is.present[is.present == FALSE])
  if(num.false > 0){
    print(warning("Marker genes not present in dataset of interest. Possibly wrongly named"))
  }
  
  #Annotate clusters using reference dataset
  counts(reference) <- logcounts(reference)
  rownames(reference) <- rowData(reference)[,symbol]
  available <- intersect(rownames(reference), rownames(sce))
  pred_labels <- SingleR(test=sce[available,], ref=reference[available,], restrict = genes, 
                         labels=colData(reference)[,annotation], de.method="wilcox")
  return(pred_labels$pruned.labels)
}

# run
anno.comHVGs <- annotate_clust(sce = sce, reference = reference, genes = common.HVGs, symbol = "Symbol", annotation = "annotation")
sce_comb$CommonHVGs_spitz <- anno.comHVGs

df <- as.data.frame(reducedDim(sce_comb, "UMAP"))
df$SingleR_spitz <- sce_comb$CommonHVGs_spitz

samp1 <- sample(1:nrow(df))
plot <- ggplot(df[samp1,], aes(x=UMAP1[samp1], y=UMAP2[samp1])) + 
      geom_point(col="black", size=0.9) +
      geom_point(col="white", size=0.5) +
      geom_point(aes(col=SingleR_spitz[samp1]), size=0.455, alpha=0.75) + scale_color_manual(values=colours_pop) +theme_void() +
      guides(colour = guide_legend(override.aes = list(alpha = 1, size=3)))
```

# Confusion matrix - Jaccard Similarity 

Compare assigned cell type annotations between the two datasets

```{r}
drummond <- as.character(sce_comb$annotation)
spitzer <- as.character(sce_comb$CommonHVGs_spitz)

calculate_jaccard <- function(x, y) {
  intersection <- length(intersect(x, y))
  union <- length(union(x, y))
  jaccard_index <- intersection / union
  return(jaccard_index)
}

unique_drummond <- unique(drummond)
unique_spitzer <- unique(spitzer)

jaccard_matrix <- matrix(nrow = length(unique_drummond), ncol = length(unique_spitzer), dimnames = list(unique_drummond, unique_spitzer))

for (i in seq_along(unique_drummond)) {
  for (j in seq_along(unique_spitzer)) {
    cells_in_i <- which(drummond == unique_drummond[i])
    cells_in_j <- which(spitzer == unique_spitzer[j])
    jaccard_matrix[i, j] <- calculate_jaccard(cells_in_i, cells_in_j)
  }
}
jaccard_matrix

jaccard_matrix <- jaccard_matrix[, -8]

p <- pheatmap(jaccard_matrix,
         color = colorRampPalette(c("#fffee0", "orange", "red"))(120),
         display_numbers = T,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean")

jaccard_matrix <- jaccard_matrix[, c("AC", "OC")]
jaccard_matrix <- jaccard_matrix[c("AC-like", "OPC-like"), ]

p <- pheatmap(jaccard_matrix,
         color = colorRampPalette(c("#fffee0", "orange", "red"))(120),
         display_numbers = T,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean")
```

## Rank-Rank Hypergeometric Overlap analysis

* GSEA pathways
* scores: -log10(pvalue) * sign(effectSize)
* Checked that both studies have performed GSEA in the same direction (Pre-treatment vs Post-treatment)

```{r}
# Spitzer et al. 2024 downloaded GSEA results
gsea_spitz <- read_xlsx("/public_datasets/spitzer/mmc4 (1).xlsx", sheet = 2, skip = 2)
unique_sets <- gsea_spitz[gsea_spitz$`Associated program` %in% c("AC-cancer", "AC-normal", "OC-normal", "OPC-normal", "Cell-cycle", "NPC-cancer", "OC-cancer"), ]

# average scores for certain modules
sum_sets <- unique_sets %>%
  dplyr::group_by(`Associated program`) %>%
  dplyr::summarise(
    `BH-adjusted p value` = mean(`BH-adjusted p value`, na.rm = TRUE),
    NES = mean(NES, na.rm = TRUE),
    pval = mean(pval, na.rm = TRUE),
    `Geneset size` = mean(`Geneset size`, na.rm = TRUE),
    `Fraction in ranked list edges` = mean(`Fraction in ranked list edges`, na.rm = TRUE),
    Pathway = dplyr::first(`Associated program`)  
  )
sum_sets$Pathway <- c("AC-like", "Astrocyte", "Cell-cycle", "NPC-like", "OPC-like", "Oligodendrocyte", "OPC-normal")

gsea_spitz <- rbind(sum_sets, gsea_spitz)
colnames(gsea_spitz) <- c("Associated Program", "padj", "NES", "pval", "size", "ES", "pathway")

# GSEA all cell type populations (Spitzer GSEA method)
gsea_drummond <- readRDS("output/gsea/gsea_results_all.rds")
gsea_drummond <- gsea_drummond$GSEA_Results
```

```{r}
# Rank-Rank Hypergeometric Overlap analysis
drummond_df <- gsea_drummond %>%
  dplyr::mutate(score = -log10(padj) * sign(NES)) %>%
  dplyr::select(pathway, score) %>%
  dplyr::filter(!is.na(score))

spitzer_df <- gsea_spitz %>%
  dplyr::mutate(score = -log10(padj) * sign(NES)) %>%
  dplyr::select(pathway, score) %>%
  dplyr::filter(!is.na(score))

paths <- intersect(drummond_df$pathway, spitzer_df$pathway)
drummond_df <- drummond_df[drummond_df$pathway %in% paths, ]
spitzer_df <- spitzer_df[spitzer_df$pathway %in% paths, ]
drummond_df$pathway <- factor(drummond_df$pathway, levels = paths)
spitzer_df$pathway <- factor(spitzer_df$pathway, levels = paths)
drummond_df <- drummond_df[order(drummond_df$pathway), ]
spitzer_df <- spitzer_df[order(spitzer_df$pathway), ]
drummond_df$pathway <- as.character(drummond_df$pathway)
spitzer_df$pathway <- as.character(spitzer_df$pathway)

RRHO_obj <-  RRHO2_initialize(as.data.frame(drummond_df), as.data.frame(spitzer_df), labels = c("Drummond", "Spitzer"), log10.ind=TRUE)
RRHO2_heatmap(RRHO_obj)

RRHO2_vennDiagram(RRHO_obj, type="dd")
RRHO2_vennDiagram(RRHO_obj, type="uu")
RRHO2_vennDiagram(RRHO_obj, type="ud")
RRHO2_vennDiagram(RRHO_obj, type="du")
```

