---
title: "Spitzer et al. 2024 comparison of snRNA-seq data"
output: html_document
---

The following code relates to the comparison of snRNA-seq data to data published in a similar study by Spitzer et al., 2024 (https://www.sciencedirect.com/science/article/pii/S153561082400093X?dgcid=author#sec4)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SummarizedExperiment)
library(HDF5Array)
library(SingleCellExperiment)
library(intrinsicDimension)
library(scater)
library(scuttle)
library(BiocParallel)
library(destiny)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(scran)
library(limma)
library(ggrepel)
library(tidyverse)
library(reshape2)
library(BiocParallel)
library(bluster)
library(fgsea)
library(pheatmap)
library(MASS)
library(lme4)
library(edgeR)
library(zoo)
library(Seurat)
library(GOSemSim)
library(RRHO2)

colours_comb <- c("#F28482", "#9D4EDD", "#6A994E", "#3A7CA5", "#EF233C",
               "#F9C8C7", "#CDDEB9", "#A7C957", "#63A2C8", "#000000", "#FFEA00", "#F4F3EE",
               "#FAA307", "#FFBA08", "#FFD60A", "#E85D04", "#F48C06", "#F2CC8F",
               "#656D4A", "#A4AC86", "#B6AD90", "#7F4F24", "#A68A64", "#D6D1C1",
               "#000000", "#ADB5BD",
               "#9D0208")
names(colours_comb) <- c("AC-like", "MES-like", "OPC-like", "NPC-like", "Progenitor",
                      "Astrocyte", "Oligodendrocyte", "OPC", "Neuron", "Vasculature", "Immune", "Unknown",
                      "T Cell", "NK Cell", "B Cell", "CD8 T Cell", "CD4 T Cell", "T Reg Cell", 
                      "Dendritic Cell", "Neutrophil", "Monocyte", "Microglia", "Macrophage", "Mast Cell",
                      "Endothelial", "Mural",
                      "Death")

colours_pop <- c("#F28482", "grey", "#000000", "#FFEA00", "#3A7CA5", "#6A994E", "#CDDEB9", "#FF474C", "#F4F3EE")
names(colours_pop) <- c("AC", "Doublet", "Endothel", "Macrophage", "NPC", "OC", "Oligodendrocyte", "Undifferentiated", "Unresolved")

colours_sample <- c("#0C7BDC", "#E66100")
names(colours_sample) <- c("Pre-treatment", "Post-treatment")
```

# Annotate snRNA-seq data using Spitzer dataset as the reference

Automatic annotation run via SingleR

* Common HVGs between snRNA-seq data and Spitzer et al. reference (matched Oligodendroglioma dataset) used for SingleR scoring 

```{r}
sce_comb <- loadHDF5SummarizedExperiment("/data/process/", prefix="sce_integrated_")
# remove participant with low tumor purity
sce_comb <- sce_comb[, !sce_comb$Patient == "A-04"]
sce_comb$Patient <- factor(sce_comb$Patient)

# Spitzer SingleR results 
reference <- readRDS("/public_datasets/spitzer/GSE260928_SNRNA/matched_oligo_sce.rds")
# gene inputs
TopHVGs <- function(x){
  dec.parse <- modelGeneVar(x)
  TopHVGs <- getTopHVGs(dec.parse, n=5000)
  return(TopHVGs)
}
top.HVGs <- TopHVGs(sce)

ReferenceHVGs <- function(x, symbol="feature_name"){
  rownames(x) <- rowData(x)[,symbol]
  dec <- modelGeneVar(x)
  Reference_HVGs <- getTopHVGs(dec, n=5000)
  return(Reference_HVGs)
}
reference.HVGs <- ReferenceHVGs(reference, symbol = "Symbol")

CommonHVGs <- function(TopHVGs, ReferenceHVGs){
  CommonHVGs <- intersect(TopHVGs, ReferenceHVGs)
  return(CommonHVGs)
}
common.HVGs <- CommonHVGs(TopHVGs = top.HVGs, ReferenceHVGs = reference.HVGs)
length(common.HVGs)

# function
annotate_clust <- function(sce, reference, genes, symbol="feature_name", annotation="annotation_level_3"){
  
  #Check for excessive amount of Ensembl gene names 
  is.ens <- grepl("^ENSG", genes, ignore.case = TRUE)
  num.true <- length(is.ens[is.ens == TRUE])
  if(num.true == length(genes)){
    print(warning("Ensembl genes detected in gene names"))
  }
  
  #Check Marker Genes are actually present in dataset of interest
  is.present <- genes %in% rownames(sce)
  num.false <- length(is.present[is.present == FALSE])
  if(num.false > 0){
    print(warning("Marker genes not present in dataset of interest. Possibly wrongly named"))
  }
  
  #Annotate clusters using reference dataset
  counts(reference) <- logcounts(reference)
  rownames(reference) <- rowData(reference)[,symbol]
  available <- intersect(rownames(reference), rownames(sce))
  pred_labels <- SingleR(test=sce[available,], ref=reference[available,], restrict = genes, 
                         labels=colData(reference)[,annotation], de.method="wilcox")
  return(pred_labels$pruned.labels)
}

# run
anno.comHVGs <- annotate_clust(sce = sce, reference = reference, genes = common.HVGs, symbol = "Symbol", annotation = "annotation")
sce_comb$CommonHVGs_spitz <- anno.comHVGs

df <- as.data.frame(reducedDim(sce_comb, "UMAP"))
df$SingleR_spitz <- sce_comb$CommonHVGs_spitz

samp1 <- sample(1:nrow(df))
plot <- ggplot(df[samp1,], aes(x=UMAP1[samp1], y=UMAP2[samp1])) + 
      geom_point(col="black", size=0.9) +
      geom_point(col="white", size=0.5) +
      geom_point(aes(col=SingleR_spitz[samp1]), size=0.455, alpha=0.75) + scale_color_manual(values=colours_pop) +theme_void() +
      guides(colour = guide_legend(override.aes = list(alpha = 1, size=3)))
```

# Confusion matrix - Jaccard Similarity 

Compare assigned cell type annotations between the two datasets

```{r}
drummond <- as.character(sce_comb$annotation)
spitzer <- as.character(sce_comb$CommonHVGs_spitz)

calculate_jaccard <- function(x, y) {
  intersection <- length(intersect(x, y))
  union <- length(union(x, y))
  jaccard_index <- intersection / union
  return(jaccard_index)
}

unique_drummond <- unique(drummond)
unique_spitzer <- unique(spitzer)

jaccard_matrix <- matrix(nrow = length(unique_drummond), ncol = length(unique_spitzer), dimnames = list(unique_drummond, unique_spitzer))

for (i in seq_along(unique_drummond)) {
  for (j in seq_along(unique_spitzer)) {
    cells_in_i <- which(drummond == unique_drummond[i])
    cells_in_j <- which(spitzer == unique_spitzer[j])
    jaccard_matrix[i, j] <- calculate_jaccard(cells_in_i, cells_in_j)
  }
}
jaccard_matrix

jaccard_matrix <- jaccard_matrix[, -8]

p <- pheatmap(jaccard_matrix,
         color = colorRampPalette(c("#fffee0", "orange", "red"))(120),
         display_numbers = T,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean")

jaccard_matrix <- jaccard_matrix[, c("AC", "OC")]
jaccard_matrix <- jaccard_matrix[c("AC-like", "OPC-like"), ]

p <- pheatmap(jaccard_matrix,
         color = colorRampPalette(c("#fffee0", "orange", "red"))(120),
         display_numbers = T,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean")
```

## Rank-Rank Hypergeometric Overlap analysis

* GSEA pathways
* scores: -log10(pvalue) * sign(effectSize)
* Checked that both studies have performed GSEA in the same direction (Pre-treatment vs Post-treatment)

```{r}
# Spitzer et al. 2024 downloaded GSEA results
gsea_spitz <- read_xlsx("/public_datasets/spitzer/mmc4 (1).xlsx", sheet = 2, skip = 2)
unique_sets <- gsea_spitz[gsea_spitz$`Associated program` %in% c("AC-cancer", "AC-normal", "OC-normal", "OPC-normal", "Cell-cycle", "NPC-cancer", "OC-cancer"), ]

# average scores for certain modules
sum_sets <- unique_sets %>%
  dplyr::group_by(`Associated program`) %>%
  dplyr::summarise(
    `BH-adjusted p value` = mean(`BH-adjusted p value`, na.rm = TRUE),
    NES = mean(NES, na.rm = TRUE),
    pval = mean(pval, na.rm = TRUE),
    `Geneset size` = mean(`Geneset size`, na.rm = TRUE),
    `Fraction in ranked list edges` = mean(`Fraction in ranked list edges`, na.rm = TRUE),
    Pathway = dplyr::first(`Associated program`)  
  )
sum_sets$Pathway <- c("AC-like", "Astrocyte", "Cell-cycle", "NPC-like", "OPC-like", "Oligodendrocyte", "OPC-normal")

gsea_spitz <- rbind(sum_sets, gsea_spitz)
colnames(gsea_spitz) <- c("Associated Program", "padj", "NES", "pval", "size", "ES", "pathway")

# GSEA all cell type populations (Spitzer GSEA method)
gsea_drummond <- readRDS("output/gsea/gsea_results_all.rds")
gsea_drummond <- gsea_drummond$GSEA_Results
```

```{r}
# Rank-Rank Hypergeometric Overlap analysis
drummond_df <- gsea_drummond %>%
  dplyr::mutate(score = -log10(padj) * sign(NES)) %>%
  dplyr::select(pathway, score) %>%
  dplyr::filter(!is.na(score))

spitzer_df <- gsea_spitz %>%
  dplyr::mutate(score = -log10(padj) * sign(NES)) %>%
  dplyr::select(pathway, score) %>%
  dplyr::filter(!is.na(score))

paths <- intersect(drummond_df$pathway, spitzer_df$pathway)
drummond_df <- drummond_df[drummond_df$pathway %in% paths, ]
spitzer_df <- spitzer_df[spitzer_df$pathway %in% paths, ]
drummond_df$pathway <- factor(drummond_df$pathway, levels = paths)
spitzer_df$pathway <- factor(spitzer_df$pathway, levels = paths)
drummond_df <- drummond_df[order(drummond_df$pathway), ]
spitzer_df <- spitzer_df[order(spitzer_df$pathway), ]
drummond_df$pathway <- as.character(drummond_df$pathway)
spitzer_df$pathway <- as.character(spitzer_df$pathway)

RRHO_obj <-  RRHO2_initialize(as.data.frame(drummond_df), as.data.frame(spitzer_df), labels = c("Drummond", "Spitzer"), log10.ind=TRUE)
RRHO2_heatmap(RRHO_obj)

RRHO2_vennDiagram(RRHO_obj, type="dd")
RRHO2_vennDiagram(RRHO_obj, type="uu")
RRHO2_vennDiagram(RRHO_obj, type="ud")
RRHO2_vennDiagram(RRHO_obj, type="du")
```


## Responder vs Non-responder analysis

# Annotate Spitzer matched datasets

Responder:

Spitzer Matched Oligodendroglioma dataset
* 1 x matched Oligodendroglioma patient (BWH445_Pre & BWH445_Post)

```{r}
# load
sce <- readRDS("/public_datasets/spitzer/GSE260928_SNRNA/matched_oligo_sce.rds")
sce$Treatment <- ifelse(grepl("PRE", sce$Tumor), "Pre-treatment", 
                       ifelse(grepl("POST", sce$Tumor), "Post-treatment", NA))
sce$Patient <- "BWH445"
```

```{r}
# scVI batch correction performed (adapted from scvi_integration.py & snrna-seq_integration_and_annotation.Rmd scripts)
# scvi.model.SCVI.setup_anndata(adata, batch_key="Treatment")

set.seed(0010101010)
Clusters <- scran::clusterCells(sce, use.dimred="X_scVI", 
                                BLUSPARAM=SNNGraphParam(k=35, type="rank", cluster.fun="leiden"))
table(Clusters)
sce$Cluster <- Clusters

df <- reducedDim(sce, "UMAP")
df <- as.data.frame(df)
df$Sample <- sce$Treatment
df$Cluster <- sce$Cluster
df$anno <- sce$annotation
colnames(df)[1:2] <- c("UMAP1", "UMAP2")

ggplot(df[sample(1:nrow(df)),], aes(x=UMAP1, y=UMAP2, color=Sample)) + geom_point(alpha=0.5) + theme_classic()

colours_clusters <- Polychrome::palette36.colors(12)
names(colours_clusters) <- 1:12

ggplot(df[sample(1:nrow(df)),], aes(x=UMAP1, y=UMAP2, color=Cluster)) + geom_point(alpha=0.5) + 
  theme_classic() +
guides(colour = guide_legend(override.aes = list(alpha = 1, size=3))) + scale_color_manual(values=colours_clusters)

# inspect marker genes
markers <- function(x, marker){
  x$marker <- as.vector(logcounts(x[marker,]))
  plot_data <- data.frame(UMAP1 = reducedDim(x, "UMAP")[,1],
                          UMAP2 = reducedDim(x, "UMAP")[,2],
                          Expression = x$marker)
  plot_data <- plot_data[order(plot_data$Expression), ]
  plot <- ggplot(plot_data, aes(x = UMAP1, y = UMAP2, color = Expression)) +
    geom_point(size=0.5) + 
    theme(aspect.ratio = 1) + 
    ggtitle(marker) + 
    scale_color_viridis_c() +
      theme_classic()
  return(plot)
}
```

### AC-like

```{r, echo=FALSE}
marker_genes <- c("APOE","AQP4", "MGAT4C", "SPARCL1", "GFAP", "ALDOC")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### MES-like

```{r, echo=FALSE}
marker_genes <- c("PDGFRA", "CHI3L1", "TNC", "ITGB1", "NRN1", "TIMP1", "S100A10")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### OPC-like

```{r, echo=FALSE}
marker_genes <- c("APOD", "OLIG1", "OLIG2", "DLL3", "SOX8", "PDGFRA")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### Progenitor

```{r, echo=FALSE}
marker_genes <- c("TOP2A", "MKI67", "CENPF", "TUBA1B")
lapply(marker_genes, function(marker) markers(sce, marker))

marker_genes <- c("TOP2A", "MKI67", "CENPF", "TUBA1B")
lapply(marker_genes, function(marker) markers(sce_responder, marker))
```

### NPC-like

```{r, echo=FALSE}
marker_genes <- c("HES6", "STMN2", "LMO4", "RAB3C", "S100A1", "DLX5", "ELAVL4", "SOX4", "CCND2", "RBM6", "SET", "TRA2A")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### Neuron

```{r}
marker_genes <- c("NPTXR", "STMN2", "CDH13", "THY1", "LMO4")
lapply(marker_genes, function(marker) markers(sce, marker))

# excitatory
marker_genes <- c("NRN1", "HOPX", "HS3ST4", "TRPM3", "EPHA4", "LMO4")
lapply(marker_genes, function(marker) markers(sce, marker))

# inhibitory
marker_genes <- c("MAF", "GRIK1", "SPOCK3", "KIT")
lapply(marker_genes, function(marker) markers(sce, marker))

# interneuron
marker_genes <- c("PVALB", "VIP", "CCK", "CRH", "HTR3A", "GAD1", "GAD2")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### OPC

```{r}
marker_genes <- c("CNP", "CA10",  "APOD", "GPNMB", "CSPG4", "CD9")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### Endothelial

```{r, echo=FALSE}
marker_genes <- c("IGFBP7", "DCN", "COL4A2")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### Astrocyte

```{r, echo=FALSE}
marker_genes <- c("MGST1", "NTRK2", "SLC1A3", "SLC1A2", "FAM107A", "CST3", "GJA1")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### Oligodendrocyte

```{r, echo=FALSE}
marker_genes <- c("ERMN", "MBP", "MAL", "MOG")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### Immune Markers

```{r, echo=FALSE}
# Immune Markers

## T Cell
marker_genes <- c("CD4", "CD8A", "CD27", "PTPRC", "CD28", "KLRB1", "CD69", "FOXP3")
lapply(marker_genes, function(marker) markers(sce, marker))
## B-cells
marker_genes <- c("CD79A", "BANK1")
lapply(marker_genes, function(marker) markers(sce, marker))
## Macrophage
marker_genes <- c("CD163", "CD68")
lapply(marker_genes, function(marker) markers(sce, marker))
## Neutrophil
marker_genes <- c("ITGAX", "SLC11A1")
lapply(marker_genes, function(marker) markers(sce, marker))
## Monocyte
marker_genes <- c("VCAN", "FCER1G", "CD14")
lapply(marker_genes, function(marker) markers(sce, marker))
## Dendritic Cell
marker_genes <- c("S100A4", "HLA-DQA1", "HLA-DQB1")
lapply(marker_genes, function(marker) markers(sce, marker))
## NK cells
marker_genes <- c("KLRC1", "GNLY") 
lapply(marker_genes, function(marker) markers(sce, marker))
## Microglia
marker_genes <- c("P2RY12", "CX3CR1")
lapply(marker_genes, function(marker) markers(sce, marker))
```

```{r}
# assign final annotations
annotation <- read.csv(file = "/data/process/spitzer_annos_oligo.csv")
sce$annotation <- annotation$annotation[match(sce$Cluster, annotation$cluster)]
sce$annotation <- factor(sce$annotation)
df$annotation <- sce$annotation

# Plot UMAP
sce_responder <- sce

df <- as.data.frame(reducedDim(sce_responder, "UMAP"))
colnames(df) <- c("UMAP1", "UMAP2")
df$sample <- sce_responder$Patient
df$Sample <- sce_responder$Treatment
df$annotation <- sce_responder$annotation

samp1 <- sample(1:nrow(df))
themes <- theme_void()

plot <- ggplot(df[samp1,], aes(x=UMAP1[samp1], y=UMAP2[samp1])) + 
      geom_point(col="black", size=0.9) +
      geom_point(col="white", size=0.5) +
      geom_point(aes(col=annotation[samp1]), size=0.455, alpha=0.75) +
      scale_color_manual(values=colours_comb) + themes +
      guides(colour = guide_legend(override.aes = list(alpha = 1, size=3)))
```


Non-responder:

Spitzer Matched Astrocytoma dataset
* 2 x matched Astrocytoma patients (BWH5033_Pre, BWH5033_Post & PARIS_Pre, PARIS_Post)
* Spitzer Matched Astrocytoma datasets combined: 27690 genes, 19363 cells

```{r}
# load UMI
spitzer1 <- read.table("/public_datasets/spitzer/GSE260928_SNRNA/GSE260928_Chromium_Astrocytoma1_processed_UMI.tsv")
spitzer2 <- read.table("/public_datasets/spitzer/GSE260928_SNRNA/GSE260928_Chromium_Astrocytoma2_processed_UMI.tsv")
same <- intersect(rownames(spitzer1), rownames(spitzer2))
spitzer1 <- spitzer1[same, ]
spitzer2 <- spitzer2[same, ]
spitzer <- cbind(spitzer1, spitzer2)
rm(spitzer1, spitzer2)
gc()

matched <- data.frame(cell_id = colnames(spitzer))
rownames(matched) <- matched$cell_id
matched$Patient <- sub("_.*", "", matched$cell_id)
matched$Treatment <- ifelse(grepl("PRE", matched$cell_id), "Pre-treatment", 
                       ifelse(grepl("POST", matched$cell_id), "Post-treatment", NA))
```

```{r}
# convert to SCE object
sce <- SingleCellExperiment(
  assays = list(counts = as.matrix(spitzer)),
  colData = matched
)

# logcounts
sce <- logNormCounts(sce)
sce$Treatment <- factor(sce$Treatment)

sce$sample <- paste0(sce$Patient, "_", sce$Treatment)

# scVI batch correction performed (adapted from scvi_integration.py & snrna-seq_integration_and_annotation.Rmd scripts)
# scvi.model.SCVI.setup_anndata(adata, batch_key="sample")

set.seed(0010101010)
Clusters <- scran::clusterCells(sce, use.dimred="X_scVI", 
                                BLUSPARAM=SNNGraphParam(k=50, type="rank", cluster.fun="leiden"))
table(Clusters)
sce$Cluster <- Clusters

df <- reducedDim(sce, "UMAP")
df <- as.data.frame(df)
df$Sample <- sce$Treatment
df$Cluster <- sce$Cluster
df$sample <- sce$sample
colnames(df)[1:2] <- c("UMAP1", "UMAP2")

ggplot(df[sample(1:nrow(df)),], aes(x=UMAP1, y=UMAP2, color=Sample)) + geom_point(alpha=0.5) + theme_classic()
ggplot(df[sample(1:nrow(df)),], aes(x=UMAP1, y=UMAP2, color=sample)) + geom_point(alpha=0.5) + theme_classic()

colours_clusters <- Polychrome::palette36.colors(12)
names(colours_clusters) <- 1:12

ggplot(df[sample(1:nrow(df)),], aes(x=UMAP1, y=UMAP2, color=Cluster)) + geom_point(alpha=0.5) + 
  theme_classic() +
guides(colour = guide_legend(override.aes = list(alpha = 1, size=3))) + scale_color_manual(values=colours_clusters)
```

### AC-like

```{r, echo=FALSE}
marker_genes <- c("APOE","AQP4", "MGAT4C", "SPARCL1", "GFAP", "ALDOC")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### MES-like

```{r, echo=FALSE}
marker_genes <- c("PDGFRA", "CHI3L1", "TNC", "ITGB1", "NRN1", "TIMP1", "S100A10")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### OPC-like

```{r, echo=FALSE}
marker_genes <- c("APOD", "OLIG1", "OLIG2", "DLL3", "SOX8", "PDGFRA")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### Progenitor

```{r, echo=FALSE}
marker_genes <- c("TOP2A", "MKI67", "CENPF", "TUBA1B")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### NPC-like

```{r, echo=FALSE}
marker_genes <- c("HES6", "STMN2", "LMO4", "RAB3C", "S100A1", "DLX5", "ELAVL4", "SOX4", "CCND2", "RBM6", "SET", "TRA2A")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### Neuron

```{r}
marker_genes <- c("NPTXR", "STMN2", "CDH13", "THY1", "LMO4")
lapply(marker_genes, function(marker) markers(sce, marker))

# excitatory
marker_genes <- c("NRN1", "HOPX", "HS3ST4", "TRPM3", "EPHA4", "LMO4")
lapply(marker_genes, function(marker) markers(sce, marker))

# inhibitory
marker_genes <- c("MAF", "GRIK1", "SPOCK3", "KIT")
lapply(marker_genes, function(marker) markers(sce, marker))

# interneuron
marker_genes <- c("PVALB", "VIP", "CCK", "CRH", "HTR3A", "GAD1", "GAD2")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### OPC

```{r}
marker_genes <- c("CNP", "CA10",  "APOD", "GPNMB", "CSPG4", "CD9")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### Endothelial

```{r, echo=FALSE}
marker_genes <- c("IGFBP7", "DCN", "COL4A2")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### Astrocyte

```{r, echo=FALSE}
marker_genes <- c("MGST1", "NTRK2", "SLC1A3", "SLC1A2", "FAM107A", "CST3", "GJA1")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### Oligodendrocyte

```{r, echo=FALSE}
marker_genes <- c("ERMN", "MBP", "MAL", "MOG")
lapply(marker_genes, function(marker) markers(sce, marker))
```

### Immune Markers

```{r, echo=FALSE}
# Immune Markers

## T Cell
marker_genes <- c("CD4", "CD8A", "CD27", "PTPRC", "CD28", "KLRB1", "CD69", "FOXP3")
lapply(marker_genes, function(marker) markers(sce, marker))
## B-cells
marker_genes <- c("CD79A", "BANK1")
lapply(marker_genes, function(marker) markers(sce, marker))
## Macrophage
marker_genes <- c("CD163", "CD68")
lapply(marker_genes, function(marker) markers(sce, marker))
## Neutrophil
marker_genes <- c("ITGAX", "SLC11A1")
lapply(marker_genes, function(marker) markers(sce, marker))
## Monocyte
marker_genes <- c("VCAN", "FCER1G", "CD14")
lapply(marker_genes, function(marker) markers(sce, marker))
## Dendritic Cell
marker_genes <- c("S100A4", "HLA-DQA1", "HLA-DQB1")
lapply(marker_genes, function(marker) markers(sce, marker))
## NK cells
marker_genes <- c("KLRC1", "GNLY") 
lapply(marker_genes, function(marker) markers(sce, marker))
## Microglia
marker_genes <- c("P2RY12", "CX3CR1")
lapply(marker_genes, function(marker) markers(sce, marker))
```

```{r}
# assign final annotations
annotation <- read.csv(file = "/data/process/spitzer_annos.csv")
sce$annotation <- annotation$annotation[match(sce$Cluster, annotation$cluster)]
sce$annotation <- factor(sce$annotation)
df$annotation <- sce$annotation

# Plot UMAP - PARIS only
sce <- sce[, sce$Patient == "PARIS"]

df <- as.data.frame(reducedDim(sce, "UMAP"))
colnames(df) <- c("UMAP1", "UMAP2")
df$sample <- sce$Patient
df$annotation <- sce$annotation

samp1 <- sample(1:nrow(df))
themes <- theme_void()

plot <- ggplot(df[samp1,], aes(x=UMAP1[samp1], y=UMAP2[samp1])) + 
      geom_point(col="black", size=0.9) +
      geom_point(col="white", size=0.5) +
      geom_point(aes(col=annotation[samp1]), size=0.455, alpha=0.75) +
      scale_color_manual(values=colours_comb) + themes +
      guides(colour = guide_legend(override.aes = list(alpha = 1, size=3)))
```

# LOESS plot

```{r}
# per cell population
calculate_logfc <- function(sce, anno, Treatment = "Treatment", condition1 = "Pre-treatment", condition2 = "Post-treatment") {
  # cell population
  sce_info <- sce[, sce$annotation == anno]
  expr_condition1 <- assay(sce_info[, sce_info[[Treatment]] == condition1], "logcounts")
  expr_condition2 <- assay(sce_info[, sce_info[[Treatment]] == condition2], "logcounts")
  logfc <- rowMeans(expr_condition2) - rowMeans(expr_condition1)
  return(logfc)
}

# non-responders
sce_nonresponder <- sce
sce_nonresponder <- sce_nonresponder[, sce_nonresponder$annotation %in% c("AC-like", "OPC-like", "Progenitor")]
sce_nonresponder$annotation <- factor(sce_nonresponder$annotation, levels = c("AC-like", "OPC-like", "Progenitor"))
# patient PARIS only
sce_nonresponder <- sce_nonresponder[, sce_nonresponder$Patient == "PARIS"]

anno_values <- unique(sce_nonresponder$annotation)
logfc_results_nonresponder <- lapply(anno_values, function(anno) {
  calculate_logfc(sce_nonresponder, anno = anno)
})
logfc_results_nonresponder <- do.call(cbind, logfc_results_nonresponder)
colnames(logfc_results_nonresponder) <- anno_values

rm(sce_nonresponder)
gc()

# responder
sce_responder <- sce_responder[, sce_responder$annotation %in% c("AC-like", "OPC-like", "Progenitor")]
sce_responder$annotation <- factor(sce_responder$annotation, levels = c("AC-like", "OPC-like", "Progenitor"))

anno_values <- unique(sce_responder$annotation)
logfc_results_responder <- lapply(anno_values, function(anno) {
  calculate_logfc(sce_responder, anno = anno)
})
logfc_results_responder <- do.call(cbind, logfc_results_responder)
colnames(logfc_results_responder) <- c("Progenitor", "AC-like", "OPC-like")

colnames(logfc_results_responder) <- paste0(colnames(logfc_results_responder), "_Responder")
colnames(logfc_results_nonresponder) <- paste0(colnames(logfc_results_nonresponder), "_Non-Responders")

keep <- intersect(rownames(logfc_results_responder), rownames(logfc_results_nonresponder))
logfc_results_responder <- logfc_results_responder[keep, ]
logfc_results_nonresponder <- logfc_results_nonresponder[keep, ]

logfc <- cbind(logfc_results_responder, logfc_results_nonresponder)

# Difference in logFC (+ve enriched in Responder, -ve enriched in Non-Responders)
anno_values <- c("AC-like", "OPC-like", "Progenitor")
logfc_difference <- list()
for (i in anno_values) {
  responder_col <- paste0(i, "_Responder")
  nonresponder_col <- paste0(i, "_Non-Responders")
  
  logfc_difference[[i]] <- logfc[, responder_col] - logfc[, nonresponder_col]
}

ranked_list <- list()
for (i in names(logfc_difference)) {
  ranked_list[[i]] <- sort(logfc_difference[[i]], decreasing = TRUE)
}

for(i in names(ranked_list)){
    gsea_subset <- ranked_list[[i]]
    stat <- as.data.frame(gsea_subset)
    stat <- cbind(rownames(stat), stat)
    colnames(stat) <- c("gene", "stat")
    
    stat <- merge(stat, program_data, by = "gene", all.x = TRUE)
    stat <- stat[order(stat$stat),]  # Sort by stat
    
    # Define the range of stat
    min_stat <- min(stat$stat)
    max_stat <- max(stat$stat)

    # Calculate the window range
    step_size <- 0.10
    window_size <- 30  # Approximate number of genes per window

    # Create a sequence of window centers
    window_centers <- seq(min_stat, max_stat, by = step_size)
    
    # Create windows around each center to contain approximately 30 genes
    window_limits <- lapply(window_centers, function(center) {
        lower_bound <- center - (step_size / 2)
        upper_bound <- center + (step_size / 2)
        close_genes <- stat$stat >= lower_bound & stat$stat < upper_bound
        # Adjust bounds if less than 30 genes are found
        while(sum(close_genes) < window_size && (lower_bound > min_stat || upper_bound < max_stat)) {
            if (lower_bound > min_stat) lower_bound <- lower_bound - step_size / 4
            if (upper_bound < max_stat) upper_bound <- upper_bound + step_size / 4
            close_genes <- stat$stat >= lower_bound & stat$stat < upper_bound
        }
        list(start = lower_bound, end = upper_bound)
    })

    # percentages for each window
    calc_percentages <- function(limits) {
        subset <- stat %>%
            filter(stat >= limits$start, stat < limits$end, !is.na(Program)) 
        
        total_genes <- nrow(subset)
        all_programs <- unique(program_data$Program)
        
        if(total_genes == 0) {
            return(data.frame(Program = all_programs, Percentage = 0, Window = (limits$start + limits$end) / 2))
        }

        percentages <- subset %>%
            dplyr::group_by(Program) %>%
            dplyr::summarise(Percentage = n() / total_genes * 100, .groups = 'drop')
        
        complete_percentages <- merge(data.frame(Program = all_programs), percentages, by = "Program", all.x = TRUE)
        complete_percentages[is.na(complete_percentages$Percentage), "Percentage"] <- 0
        complete_percentages$Window <- (limits$start + limits$end) / 2
        
        return(complete_percentages)
    }
    stat <- na.omit(stat)

    percentages <- lapply(window_limits, calc_percentages)
    percentages_df <- do.call(rbind, percentages)

    plot <- ggplot(percentages_df, aes(x = Window, y = Percentage, color = Program)) +
        geom_point(alpha = 0.5, size = 0.8) +
        geom_smooth(method = "loess", span = 0.5, se = FALSE) +
        labs(title = paste0(i),
             x = "Responder vs Non-Responders [Difference in LogFC]",
             y = "% Genes in Tumour Program [in sliding window]",
             color = "Gene Program") +
        theme_classic() + 
        scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 10)) +
        scale_color_manual(values = c("AC-like" = "#F28482", "OPC-like" = "#6A994E", "NPC-like" = "#3A7CA5", "MES-like" = "#9D4EDD"))

    print(plot)
}
```

# Progenitor ternary metabolism plots

```{r}
glycolysis_genes <- c("HK1", "HK2", "HK3", "GPI", "PFKL", "PFKM", "PFKP", 
                      "ALDOA", "ALDOB", "ALDOC", "TPI1", "GAPDH", "GAPDHS", 
                      "PGK1", "PGK2", "PGAM1", "PGAM2", "PGAM4", "ENO1", 
                      "ENO2", "ENO3", "PKM", "PKLR")

krebs_genes <- c("CS", "IDH1", "IDH2", "IDH3A", "IDH3B", "IDH3G", "OGDH", "OGDHL", "SDHA", "SDHB", "SDHC", "SDHD", "FH", "MDH1", "MDH2")

gene_sets <- readRDS("/public_datasets/gene_sets/all_genesets_list.rds")
interest <- gene_sets[grep("GOBP_LACTATE_METABOLIC_PROCESS", names(gene_sets))]
lactate_genes <- interest[[1]]
```

```{r}
# responder
sce_responder <- sce_responder[, sce_responder$annotation %in% c("Progenitor")]
sce_responder$Info <- paste0(sce_responder$Patient, "_", sce_responder$Treatment)

sce_responder$glycolysis_score <- colMeans(logcounts(sce_responder[rownames(sce_responder) %in% glycolysis_genes, ]))
sce_responder$krebs_score <- colMeans(logcounts(sce_responder[rownames(sce_responder) %in% krebs_genes, ]))
sce_responder$lactate_score <- colMeans(logcounts(sce_responder[rownames(sce_responder) %in% lactate_genes, ]))

plot_data <- data.frame(
  Patient = sce_responder$Patient,
  Info = sce_responder$Info,
  Glycolysis = sce_responder$glycolysis_score,
  Krebs = sce_responder$krebs_score,
  Lactate = sce_responder$lactate_score,
  Treatment = sce_responder$Treatment
)

# Normalise scores sum to 1 for each cell
plot_data <- transform(plot_data, Total = Glycolysis + Krebs + Lactate)
plot_data <- transform(plot_data,
                       Glycolysis = Glycolysis / Total,
                       Krebs = Krebs / Total,
                       Lactate = Lactate / Total)

means <- plot_data %>%
    dplyr::group_by(Treatment) %>%
    dplyr::summarise(
      Glycolysis = mean(Glycolysis, na.rm = TRUE),
      Krebs = mean(Krebs, na.rm = TRUE),
      Lactate = mean(Lactate, na.rm = TRUE),
      .groups = "drop"
    )

plot_data$Treatment <- factor(plot_data$Treatment, levels = c("Pre-treatment", "Post-treatment"))
plot <- ggtern(plot_data, aes(x = Glycolysis, y = Krebs, z = Lactate, fill = Treatment)) +
  stat_density_tern(aes(alpha = ..level..), geom = "polygon", color = NA) + 
  stat_density_tern(aes(color = Treatment), geom = "path", size = 0.5, alpha = 0.9) + 
    geom_point(
      data = means,
      aes(x = Glycolysis, y = Krebs, z = Lactate, fill = Treatment),
      size = 4,
      shape = 21,  
      color = "black"  
    ) +
  labs(
    x = "Glycolysis",
    y = "Krebs",
    z = "Lactate"
  ) +
  theme_classic() +
  theme_arrowlong() + 
  scale_fill_manual(values = colours_sample) +   
  scale_color_manual(values = colours_sample) +    
  scale_alpha_continuous(range = c(0.2, 0.95)) +  
  theme(
    axis.title = element_text(size = 8),       
    axis.text = element_text(size = 5),        
    strip.text = element_text(size = 10),     
    plot.title = element_text(size = 12, face = "bold")
  )

# non-responder
sce_nonresponder <- sce_nonresponder[, sce_nonresponder$annotation %in% c("Progenitor")]
sce_nonresponder$Info <- paste0(sce_nonresponder$Patient, "_", sce_nonresponder$Treatment)

sce_nonresponder$glycolysis_score <- colMeans(logcounts(sce_nonresponder[rownames(sce_nonresponder) %in% glycolysis_genes, ]))
sce_nonresponder$krebs_score <- colMeans(logcounts(sce_nonresponder[rownames(sce_nonresponder) %in% krebs_genes, ]))
sce_nonresponder$lactate_score <- colMeans(logcounts(sce_nonresponder[rownames(sce_nonresponder) %in% lactate_genes, ]))

plot_data <- data.frame(
  Patient = sce_nonresponder$Patient,
  Info = sce_nonresponder$Info,
  Glycolysis = sce_nonresponder$glycolysis_score,
  Krebs = sce_nonresponder$krebs_score,
  Lactate = sce_nonresponder$lactate_score,
  Treatment = sce_nonresponder$Treatment
)

# Normalise scores sum to 1 for each cell
plot_data <- transform(plot_data, Total = Glycolysis + Krebs + Lactate)
plot_data <- transform(plot_data,
                       Glycolysis = Glycolysis / Total,
                       Krebs = Krebs / Total,
                       Lactate = Lactate / Total)

means <- plot_data %>%
    dplyr::group_by(Treatment) %>%
    dplyr::summarise(
      Glycolysis = mean(Glycolysis, na.rm = TRUE),
      Krebs = mean(Krebs, na.rm = TRUE),
      Lactate = mean(Lactate, na.rm = TRUE),
      .groups = "drop"
    )

plot_data$Treatment <- factor(plot_data$Treatment, levels = c("Pre-treatment", "Post-treatment"))
plot <- ggtern(plot_data, aes(x = Glycolysis, y = Krebs, z = Lactate, fill = Treatment)) +
  stat_density_tern(aes(alpha = ..level..), geom = "polygon", color = NA) + 
  stat_density_tern(aes(color = Treatment), geom = "path", size = 0.5, alpha = 0.9) + 
  geom_point(
      data = means,
      aes(x = Glycolysis, y = Krebs, z = Lactate, fill = Treatment),
      size = 4,
      shape = 21, 
      color = "black" 
    ) +
  labs(
    x = "Glycolysis",
    y = "Krebs",
    z = "Lactate"
  ) +
  theme_classic() +
  theme_arrowlong() + 
  scale_fill_manual(values = colours_sample) +   
  scale_color_manual(values = colours_sample) +    
  scale_alpha_continuous(range = c(0.2, 0.95)) +  
  theme(
    axis.title = element_text(size = 8),       
    axis.text = element_text(size = 5),        
    strip.text = element_text(size = 10),     
    plot.title = element_text(size = 12, face = "bold")
  )
```

# Inspect genes of interest - Progenitors

```{r}
# DEG violin
genes_of_interest <- c("BMPER", "MYRF")

# Responder
exprs_data <- as.data.frame(t(logcounts(sce_responder)[genes_of_interest, ]))
exprs_data$Treatment <- sce_responder$Treatment
exprs_data$Treatment <- factor(exprs_data$Treatment, levels = c("Pre-treatment", "Post-treatment"))

for (i in seq_along(genes_of_interest)) {
    gene <- genes_of_interest[i]
    t_test_result <- t.test(exprs_data[[gene]] ~ exprs_data$Treatment)
    t_test_result

  gg1 <- ggplot(exprs_data, aes_string(x="Treatment", y=genes_of_interest[i], fill="Treatment")) +
    geom_violin(trim=TRUE) +
    scale_fill_manual(values=colours_sample) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.x = element_blank(),
          legend.title = element_blank()) +
    labs(
      title = paste0("p-value: ", signif(t_test_result$p.value, 3))
    )
  
  print(gg1)
}

p_values <- numeric(length(genes_of_interest))
for (i in seq_along(genes_of_interest)) {
    gene <- genes_of_interest[i]
    t_test_result <- t.test(exprs_data[[gene]] ~ exprs_data$Treatment)
    p_values[i] <- t_test_result$p.value
}

# Adjust p-values using the Benjamini-Hochberg (FDR) method
adjusted_p_values <- p.adjust(p_values, method = "BH")

results <- data.frame(
    Gene = genes_of_interest,
    Raw_p_value = p_values,
    Adjusted_p_value = adjusted_p_values
)

# Non-responder (patient PARIS only)
sce_nonresponder <- sce_nonresponder[, sce_nonresponder$Patient == "PARIS"]
exprs_data <- as.data.frame(t(logcounts(sce_nonresponder)[genes_of_interest, ]))
exprs_data$Treatment <- sce_nonresponder$Treatment
exprs_data$Treatment <- factor(exprs_data$Treatment, levels = c("Pre-treatment", "Post-treatment"))

for (i in seq_along(genes_of_interest)) {
  
  gene <- genes_of_interest[i]
  t_test_result <- t.test(exprs_data[[gene]] ~ exprs_data$Treatment)
  t_test_result
  wilcox_test_result <- wilcox.test(exprs_data[[gene]] ~ exprs_data$Treatment)
  
  gg1 <- ggplot(exprs_data, aes_string(x="Treatment", y=genes_of_interest[i], fill="Treatment")) +
    geom_violin(trim=TRUE) +
    scale_fill_manual(values=colours_sample) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          axis.title.x = element_blank(),
          legend.title = element_blank()) +
    labs(
      title = paste0("p-value: ", signif(t_test_result$p.value, 3))
    )
  
  print(gg1)
}

p_values <- numeric(length(genes_of_interest))
for (i in seq_along(genes_of_interest)) {
    gene <- genes_of_interest[i]
    t_test_result <- t.test(exprs_data[[gene]] ~ exprs_data$Treatment)
    p_values[i] <- t_test_result$p.value
}

# Adjust p-values using the Benjamini-Hochberg (FDR) method
adjusted_p_values <- p.adjust(p_values, method = "BH")

results <- data.frame(
    Gene = genes_of_interest,
    Raw_p_value = p_values,
    Adjusted_p_value = adjusted_p_values
)
```
