---
title: "Independent LGG cohort analysis (snRNA-seq data)"
output: html_document
date: "2025-01-10"
---

The following code relates to the analysis of snRNA-seq data from an independent cohort of low-grade glioma (n = 3 Oligodendroglioma, n = 8 Astrocytoma).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(SummarizedExperiment)
library(HDF5Array)
library(SingleCellExperiment)
library(intrinsicDimension)
library(scater)
library(scuttle)
library(BiocParallel)
library(SCENT)
library(destiny)
library(diptest)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(scran)
library(limma)
library(ggrepel)
library(tidyverse)
library(reshape2)
library(BiocParallel)
library(bluster)
library(fgsea)
library(pheatmap)
library(lme4)
library(edgeR)
library(zoo)
library(Seurat)
library(GOSemSim)
library(UpSetR)
library(numbat)
library(CytoTRACE2)
library(entropy)
library(reticulate)

# Load colours 
colours_comb <- c("#F28482", "#9D4EDD", "#6A994E", "#3A7CA5", "#EF233C",
               "#F9C8C7", "#CDDEB9", "#A7C957", "#63A2C8", "#10F4B9", "#d8e5e5", "#ADD8E6", "#89CFF0", "#6495ED", "#9CFDFF", "#000000", "#FFEA00", "#F4F3EE",
               "#FAA307", "#FFBA08", "#FFD60A", "#E85D04", "#F48C06", "#F2CC8F",
               "#656D4A", "#A4AC86", "#B6AD90", "#7F4F24", "#A68A64", "#D6D1C1",
               "#000000", "#ADB5BD",
               "#9D0208", "#8c0001", "#ffabff")
names(colours_comb) <- c("AC-like", "MES-like", "OPC-like", "NPC-like", "Progenitor",
                      "Astrocyte", "Oligodendrocyte", "OPC", "Neuron", "Excitatory Neuron", "PN-dev", "PN-L2/3", "PN-L4", "PN-L5/6", "Inhibitory Neuron", "Vasculature", "Immune", "Unknown",
                      "T Cell", "NK Cell", "B Cell", "CD8 T Cell", "CD4 T Cell", "T Reg Cell", 
                      "Dendritic Cell", "Neutrophil", "Monocyte", "Microglia", "Macrophage", "Mast Cell",
                      "Endothelial", "Mural",
                      "Death", "Tumor", "Normal")

colour_sample <- c("A-A"= "#3498DB", "O-A" = "#E74C3C", "A-B" = "#1ABC9C", "A-C" = "#2ECC71", "A-D" = "#9B59B6", "O-B" = "#F39C12", "O-C" = "#E67E22", "A-E" = "#34495E", "A-F" = "#5DADE2", "A-G" = "#48C9B0", "A-H" = "#85C1E9")
```

# Load dataset

```{r}
sce <- readRDS("/data/processed/sce_lgg.rds")

df <- as.data.frame(reducedDim(sce, "scvi_UMAP"))
colnames(df) <- c("UMAP1", "UMAP2")
df$sample <- sce$Sample
df$annotation <- sce$Anno2

samp1 <- sample(1:nrow(df))
themes <- theme_void()

ggplot(df[samp1,], aes(x=UMAP1[samp1], y=UMAP2[samp1])) + 
      geom_point(col="black", size=0.9) +
      geom_point(col="white", size=0.5) +
      geom_point(aes(col=annotation[samp1]), size=0.455, alpha=0.75) +
      scale_color_manual(values=colours_comb) + themes +
      guides(colour = guide_legend(override.aes = list(alpha = 1, size=3)))

glycolysis_genes <- glycolysis_genes[glycolysis_genes %in% rownames(sce)]

# remove unwanted annotations
sce <- sce[, !sce$Anno2 %in% c("Poor Quality AC-like", "Poor Quality Microglia", "Unknown")]

# append metadata
meta <- read.csv("/data/meta/lgg_meta.csv")
for (i in setdiff(colnames(meta), "X")) {
  colData(sce)[[i]] <- meta[match(sce$Sample, meta$X), i]
}

sce$Diagnosis <- ifelse(grepl("A", sce$grade), 
                         "Astrocytoma", 
                         ifelse(grepl("O", sce$grade), 
                                "Oligodendroglioma", 
                                NA))
sce$Diagnosis <- as.factor(sce$Diagnosis)
```

```{r}
# Plot UMAPs
sce$Anno2[sce$Anno2 %in% c("M1 Macrophage", "M2 Macrophage", "CD8 T Cell", "NK Cell", "Monocyte", "B Cell", "Immunosuppressive Microglia", "CD4 T Cell", "Microglia", "Neutrophil", "Activated Microglia")] <- "Immune"
sce$Anno2[sce$Anno2 %in% c("Excitatory", "CGE", "MGE")] <- "Neuron"
sce$Anno2[sce$Anno2 %in% c("Endothelial", "Mural", "Fibroblast")] <- "Vasculature"

df <- as.data.frame(reducedDim(sce, "scvi_UMAP"))
colnames(df) <- c("UMAP1", "UMAP2")
df$sample <- sce$Sample
df$annotation <- sce$Anno2

samp1 <- sample(1:nrow(df))
themes <- theme_void()

plot <- ggplot(df[samp1,], aes(x=UMAP1[samp1], y=UMAP2[samp1])) + 
      geom_point(col="black", size=0.9) +
      geom_point(col="white", size=0.5) +
      geom_point(aes(col=annotation[samp1]), size=0.455, alpha=0.75) +
      scale_color_manual(values=colours_comb) + themes +
      guides(colour = guide_legend(override.aes = list(alpha = 1, size=3)))

plot <- ggplot(df[samp1,], aes(x=UMAP1[samp1], y=UMAP2[samp1])) + 
      geom_point(col="black", size=0.9) +
      geom_point(col="white", size=0.5) +
      geom_point(aes(col=sample[samp1]), size=0.455, alpha=0.75) +
      scale_color_manual(values=colour_sample) + themes +
      guides(colour = guide_legend(override.aes = list(alpha = 1, size=3)))
```

```{r}
# now only keep cell types of interest
sce <- sce[, sce$Anno2 %in% c("AC-like", "OPC-like", "Oligodendrocyte", "Progenitor", "OPC", "Astrocyte")]
```

# Glycolysis per cell-type

```{r}
gene_sets <- readRDS("/public_datasets/gene_sets/all_genesets_list.rds")
interest <- gene_sets[grep("GOBP_GLYCOLYTIC_PROCESS", names(gene_sets))]
interest <- interest[[1]]

# top genes according to ChatGPT - enzymes that catalyze each step
glycolysis_genes <- c("HK1", "HK2", "HK3", "GPI", "PFKL", "PFKM", "PFKP", 
                      "ALDOA", "ALDOB", "ALDOC", "TPI1", "GAPDH", "GAPDHS", 
                      "PGK1", "PGK2", "PGAM1", "PGAM2", "PGAM4", "ENO1", 
                      "ENO2", "ENO3", "PKM", "PKLR")
```

```{r}
sce_glycolysis <- sce[glycolysis_genes, ]
sce$glycolysis_score <- colMeans(logcounts(sce_glycolysis))

plot_data <- data.frame(
  Anno = sce$Anno2,
  Glycolysis_Score = sce$glycolysis_score,
  Patient = sce$Sample
)

plot_data_summary <- plot_data %>%
  group_by(Anno, Patient) %>%
  summarise(Glycolysis_Score = mean(Glycolysis_Score, na.rm = TRUE), .groups = 'drop')

plot <- ggplot(plot_data_summary, aes(x = Anno, y = Glycolysis_Score, color = Patient)) +
  geom_boxplot(aes(group = Anno, fill = Anno), alpha = 0.9, outlier.shape = NA) + 
  geom_point(position = position_jitter(width = 0.28, height = 0), size = 2) +  
  labs(title = "Glycolysis",
       x = NULL,
       y = "Average Log Expression (Glycolysis)") +
  theme_classic() + 
  scale_color_manual(values = colours_comb) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_colour_manual(values = colour_sample) +
  scale_fill_manual(values = colours_comb) 
```

```{r}
# significance of glycolysis enrichment (averages across samples)
#AC-like against all others (excluding astrocyte)
plot_data_filtered <- plot_data_summary %>%
  filter(Anno != "Astrocyte")

group_A <- plot_data_filtered %>% 
  filter(Anno == "AC-like") %>% 
  pull(Glycolysis_Score)

group_Other <- plot_data_filtered %>% 
  filter(Anno != "AC-like") %>% 
  pull(Glycolysis_Score)

t_test <- t.test(group_A, group_Other, var.equal = FALSE)
adjusted_p <- p.adjust(t_test$p.value, method = "fdr")


#Astrocyte against all others (excluding AC-like)
plot_data_filtered <- plot_data_summary %>%
  filter(Anno != "AC-like")

group_A <- plot_data_filtered %>% 
  filter(Anno == "Astrocyte") %>% 
  pull(Glycolysis_Score)

group_Other <- plot_data_filtered %>% 
  filter(Anno != "Astrocyte") %>% 
  pull(Glycolysis_Score)

t_test <- t.test(group_A, group_Other, var.equal = FALSE)
adjusted_p <- p.adjust(t_test$p.value, method = "fdr")

# compare all cell types
t_test_results <- pairwise.t.test(plot_data$Glycolysis_Score, plot_data$Anno, p.adjust.method = "BH")
t_test_results
```

# KREBs per cell population

How do individual OPC-like tumor cells compare to AC-like tumor cells in their Krebs cycle expression?

```{r}
interest <- gene_sets[grep("GOBP_TRICARBOXYLIC_ACID_CYCLE", names(gene_sets))]
interest <- interest[[1]]

# top genes according to ChatGPT 
krebs_genes <- c("CS", "IDH1", "IDH2", "IDH3A", "IDH3B", "IDH3G", "OGDH", "OGDHL", "SDHA", "SDHB", "SDHC", "SDHD", "FH", "MDH1", "MDH2")
krebs_genes <- krebs_genes[krebs_genes %in% rownames(sce)]

# average log expression
sce_kreb <- sce[krebs_genes, ]
sce$krebs_score <- colMeans(logcounts(sce_kreb))

plot_data <- data.frame(
  Anno = sce$Anno2,
  Krebs_Score = sce$krebs_score,
  Patient = sce$Sample
)

plot <- ggplot(plot_data, aes(x = Anno, y = Krebs_Score, color=Anno)) +
  geom_jitter(size=0.1, alpha=0.25) + 
           geom_violin() +
  labs(title = "KREBS",
       y = "Average Log Expression (KREBS Genes)") +
  theme_classic() + scale_color_manual(values=colours_comb) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

t_test_results <- pairwise.t.test(plot_data$Krebs_Score, plot_data$Anno, p.adjust.method = "BH")
t_test_results
```

# Lactate per cell population

How do individual OPC-like tumor cells compare to AC-like tumor cells in their lactate metabolism expression?

```{r}
interest <- gene_sets[grep("GOBP_LACTATE_METABOLIC_PROCESS", names(gene_sets))]
lactate_genes <- interest[[1]]
lactate_genes <- lactate_genes[lactate_genes %in% rownames(sce)]

# average log expression
sce_lactate <- sce[lactate_genes, ]
sce$lactate_score <- colMeans(logcounts(sce_lactate))

plot_data <- data.frame(
  Anno = sce$Anno2,
  Lactate_Score = sce$lactate_score,
  Patient = sce$Sample
)

plot <- ggplot(plot_data, aes(x = Anno, y = Lactate_Score, color=Anno)) +
  geom_jitter(size=0.1, alpha=0.25) + 
           geom_violin() +
  labs(title = "Lactate",
       y = "Average Log Expression (Lactate Genes)") +
  theme_classic() + scale_color_manual(values=colours_comb) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

t_test_results <- pairwise.t.test(plot_data$Lactate_Score, plot_data$Anno, p.adjust.method = "BH")
t_test_results
```

# AC-like vs OPC-like Ternary metabolism plots

```{r}
# Krebs
interest <- gene_sets[grep("GOBP_TRICARBOXYLIC_ACID_CYCLE", names(gene_sets))]
interest <- interest[[1]]
# top genes according to ChatGPT 
krebs_genes <- c("CS", "IDH1", "IDH2", "IDH3A", "IDH3B", "IDH3G", "OGDH", "OGDHL", "SDHA", "SDHB", "SDHC", "SDHD", "FH", "MDH1", "MDH2")
krebs_genes <- krebs_genes[krebs_genes %in% rownames(sce)]

# Lactate
interest <- gene_sets[grep("GOBP_LACTATE_METABOLIC_PROCESS", names(gene_sets))]
lactate_genes <- interest[[1]]
lactate_genes <- lactate_genes[lactate_genes %in% rownames(sce)]

sce$glycolysis_score <- colMeans(logcounts(sce[glycolysis_genes, ]))
sce$krebs_score <- colMeans(logcounts(sce[krebs_genes, ]))
sce$lactate_score <- colMeans(logcounts(sce[lactate_genes, ]))

plot_data <- data.frame(
  CellType = sce$Anno2,
  Glycolysis = sce$glycolysis_score,
  Krebs = sce$krebs_score,
  Lactate = sce$lactate_score,
  Patient = sce$Sample
)

# Normalise scores sum to 1 for each cell
plot_data <- transform(plot_data, Total = Glycolysis + Krebs + Lactate)
plot_data <- transform(plot_data,
                       Glycolysis = Glycolysis / Total,
                       Krebs = Krebs / Total,
                       Lactate = Lactate / Total)

plot_data <- plot_data[plot_data$CellType %in% c("AC-like", "OPC-like"), ]

plot <- ggtern(plot_data, aes(x = Glycolysis, y = Krebs, z = Lactate, fill = CellType)) +
  stat_density_tern(aes(alpha = ..level..), geom = "polygon", color = NA) + 
  stat_density_tern(aes(color = CellType), geom = "path", size = 0.5, alpha = 0.9) + 
  labs(
    x = "Glycolysis",
    y = "Krebs",
    z = "Lactate"
  ) +
  theme_classic() +
  theme_arrowlong() + 
  scale_fill_manual(values = colours_comb) +   
  scale_color_manual(values = colours_comb) +    
  scale_alpha_continuous(range = c(0.2, 0.95)) +  
  theme(
    axis.title = element_text(size = 8),       
    axis.text = element_text(size = 5),        
    strip.text = element_text(size = 10),     
    plot.title = element_text(size = 12, face = "bold")
  )
```



